---
title: "mrpipeline User Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mrpipeline User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # run_mr()/run_coloc() narrate their progress through messages; warnings
  # stay on, since several examples below are about what they say.
  message = FALSE
)
library(mrpipeline)

# The evaluated examples need the bundled LD panel and a PLINK binary.
# genetics.binaRies::get_plink_binary() downloads PLINK on first use and
# errors if it cannot, so probe it rather than assume it. Without it the
# examples are shown with their output omitted, instead of failing the build.
bfile <- sub(
  "\\.bed$",
  "",
  system.file("extdata", "ld_ref.bed", package = "mrpipeline")
)
plink_bin <- tryCatch(
  genetics.binaRies::get_plink_binary(),
  error = function(e) ""
)
has_ld <- file.exists(paste0(bfile, ".bed")) && file.exists(plink_bin)
has_gg <- has_ld && requireNamespace("ggplot2", quietly = TRUE)
has_susie <- has_ld && requireNamespace("susieR", quietly = TRUE)
has_forestplot <- requireNamespace("forestplot", quietly = TRUE)

# Numbers quoted in the prose come from the evaluated results, never typed.
# Counts print as integers; anything else to `digits` significant figures,
# keeping trailing zeros (0.0500, not 0.05).
num <- function(x, digits = 3) {
  if (!has_ld) {
    return("[not computed]")
  }
  if (all(x == round(x))) {
    return(format(x))
  }
  formatC(x, digits = digits, format = "fg", flag = "#")
}
```

```{r no-plink-note, eval = !has_ld, echo = FALSE, results = "asis"}
cat(
  "> **Note:** PLINK or the bundled LD panel was not available when this",
  "page was built, so the examples below show their code without output.\n"
)
```

## Overview

`mrpipeline` runs two-sample Mendelian randomisation (MR) and
colocalisation for molecular exposures -- currently protein QTLs (deCODE,
UKB-PPP) and single-cell eQTLs -- against GWAS outcomes. It is built for the
drug-target question "does genetically proxied perturbation of this protein
affect this disease?", so its default mode is **cis-MR**: instruments are
drawn from a window around the gene encoding the protein.

It wraps `TwoSampleMR`, `MendelianRandomization` and `coloc` behind two
calls, `run_mr()` and `run_coloc()`, which each return a single result object
(`mr_result`, `coloc_result`) carrying the estimates, the instruments, the
harmonisation record and every diagnostic -- so a result can be inspected,
plotted and audited after the fact.

**About the examples.** Evaluated examples on this page use the data bundled
with the package: `cd40_exposure` (CD40 protein, UKB-PPP), `sjogren_outcome`
(Sjogren's disease) and a 50-SNP LD reference panel of real 1000 Genomes EUR
genotypes. Their output is produced when the page is built, not typed. The
outcome data are **partly synthetic** (49 of its 50 SNPs), so the numbers
illustrate the mechanics and are not a finding about CD40 and Sjogren's
disease. Examples that need network access or your own files are marked as
not run.

## Gene Coordinate Lookup

Use `get_gene_coords()` to programmatically retrieve gene coordinates from
Ensembl via biomaRt. This is useful for defining cis regions in `run_mr()` and
`run_coloc()` without hard-coding coordinates. These two examples query
Ensembl over the network, so they are not run here; the output shown is
representative.

```{r gene-coords, eval = FALSE}
# GRCh38 (default)
coords <- get_gene_coords(c("CD40", "APOE"))
coords
#> # A tibble: 2 x 4
#>   hgnc_symbol chromosome     start       end
#>   <chr>       <chr>          <int>     <int>
#> 1 APOE        19          44905791  44909393
#> 2 CD40        20          44746911  44758502

# GRCh37
coords_37 <- get_gene_coords("CD40", build = "grch37")
```

The returned tibble can feed directly into `run_mr()` and `run_coloc()`:

```{r use-coords, eval = FALSE}
cd40 <- get_gene_coords("CD40", build = "grch38")

mr_res <- run_mr(
  exposure = exposure_data,
  exposure_id = "CD40",
  outcome = outcome_data,
  outcome_id = "SjD",
  instrument_region = list(
    chromosome = cd40$chromosome,
    start = cd40$start,
    end = cd40$end
  )
)
```

## Controlling PLINK Resource Usage

When running many parallel jobs (e.g. on an HPC cluster), PLINK's default
behaviour of auto-detecting available threads and memory can cause problems --
each worker may try to reserve half the node's RAM. Use `plink_threads` and
`plink_memory` to cap resource usage per call.

You can set these per-call, via R options, or via environment variables
(placeholder data; not run):

```{r plink-resources, eval = FALSE}
# Per-call (highest priority)
result <- run_mr(
  exposure = exposure_data, exposure_id = "CD40",
  outcome = outcome_data, outcome_id = "SjD",
  bfile = "/path/to/ld_ref",
  plink_threads = 1,
  plink_memory = 2000
)

# Via R options (apply to all subsequent calls in the session)
options(
  mrpipeline.plink_threads = 1,
  mrpipeline.plink_memory = 2000
)

# Via environment variables (e.g. in .Renviron or a job script) -- set
# MRPIPELINE_PLINK_THREADS to 1 and MRPIPELINE_PLINK_MEMORY to 2000
```

Priority order: explicit argument > R option > environment variable > PLINK
auto-detect (`NULL`). The same parameters are available in `run_coloc()`.

## Setting Up an LD Reference Panel

Several steps need linkage disequilibrium (LD) between variants, which
`mrpipeline` computes with PLINK from a reference panel of genotypes:

| Step | Without `bfile` | With `bfile` |
|---|---|---|
| Instrument clumping in `run_mr()` | OpenGWAS API (needs an `OPENGWAS_JWT` token; rate-limited) | Local PLINK |
| `run_mr(ld_correct = TRUE)` | Error | LD matrix of the instruments |
| `run_coloc()` (SuSiE, `coloc.signals`, LD alignment) | Required -- `bfile` has no default | LD matrix of the region |
| `plot(coloc_res, type = "locuszoom")` | Required | LD colouring |

A local panel is recommended for anything beyond a quick look: it is
reproducible (the API's panel can change), it is not rate-limited, which
matters when you batch many proteins, and it is the only option for
`ld_correct` and colocalisation.

**File structure.** `bfile` is a PLINK 1 binary fileset given by its
*prefix* -- the path without an extension. `bfile = "/data/ld/EUR"` expects
`/data/ld/EUR.bed`, `/data/ld/EUR.bim` and `/data/ld/EUR.fam` to sit side by
side. Variant IDs in the `.bim` must be rsIDs matching your summary
statistics, and the panel must be on the same genome build as your data.

**Getting a panel.** Use one matching your GWAS samples' ancestry. The
MRC-IEU distributes the 1000 Genomes phase 3 panels (GRCh37, rsIDs) used by
OpenGWAS, split by super-population:

```{r ld-panel-download, eval = FALSE}
# Not run: a ~1.6 GB download.
download.file("http://fileserve.mrcieu.ac.uk/ld/1kg.v3.tgz", "1kg.v3.tgz")
untar("1kg.v3.tgz", exdir = "ld_ref")
bfile <- "ld_ref/EUR" # ld_ref/EUR.bed, ld_ref/EUR.bim, ld_ref/EUR.fam
```

The package also bundles a tiny panel -- real 1000 Genomes EUR genotypes for
503 individuals at the 50 CD40-region SNPs in `cd40_exposure` -- which the
evaluated examples on this page use:

```{r bundled-panel, eval = has_ld}
bfile <- sub(
  "\\.bed$",
  "",
  system.file("extdata", "ld_ref.bed", package = "mrpipeline")
)
basename(bfile)
```

It only covers those 50 SNPs, so it is for learning the package, not for
analysis. The script that built it, `data-raw/ld_ref.R` in the package
source, shows how to cut a panel for any region straight from the 1000
Genomes VCFs.

## Formatting GWAS Data

`run_mr()` and `run_coloc()` take an **exposure** in TwoSampleMR format
(`SNP`, `beta.exposure`, `se.exposure`, `effect_allele.exposure`,
`other_allele.exposure`, `eaf.exposure`, `pval.exposure`, ...) and an
**outcome** in the `format_gwas()` outcome schema (`rsids`, `chr`, `pos`,
`beta`, `se`, `eaf`, `pval`, `n`, `effect_allele`, `other_allele`,
`phenotype`). `format_gwas()` produces either from almost any summary
statistics file. Source-specific wrappers exist for deCODE
(`format_pqtl_decode()`), UKB-PPP (`format_pqtl_ukbppp()`) and OneK1K
single-cell eQTLs (`format_single_cell_onek1k()`); they already know their
source's allele convention, so the steps below are for everything else.

The walkthrough in this section follows a real analysis -- an
AMD outcome GWAS and a regenie exposure -- from files on disk, including a
deliberately mis-oriented file. Those files do not ship with the package, so
this code is **not run** when the page is built; the `#>` lines are
representative output, abridged where marked `...`.

### Step 1: inspect the header

Look at the column names before formatting anything:

```{r inspect-header, eval = FALSE}
path <- "genomics_data/outcome_GWAS/AMD/IAMDGC_AMD.tsv.gz"
names(data.table::fread(path, nrows = 5))
#> [1] "SNP" "CHR" "POS" "A1" "A2" "BETA" "SE" "PVALUE" "N" "FRQ" ...
```

Compare them against the alias table in `?format_gwas` (section *Column
normalisation*). Columns in the table are renamed automatically.

### Step 2: map any columns the alias table does not know

Anything not recognised needs a `col_map` entry. Other switches:
`log10_pval = TRUE` for `-log10(p)` columns, `n = ...` when the file has no
sample-size column, `bim_path` when rsIDs (or chr/pos) are missing, and
`marker_col` for compound `CHR:POS:...` identifiers.

```{r col-map, eval = FALSE}
col_map <- list(pval = "PVALUE")
```

### Step 3: decide what A1 means

This is the step that is easy to skip and expensive to get wrong. Two
conventions share the names `A1`/`A2`:

| Convention | `A1` | `A2` | `BETA` / `FRQ` refer to |
|---|---|---|---|
| PLINK, regenie, METAL, most GWAS Catalog deposits | effect allele | other allele | `A1` |
| EPACTS, RAREMETAL, VCF-derived tables | REF | ALT | `A2` |

`format_gwas()` assumes the first. A file using the second is silently read
with effect and other allele swapped, and because harmonisation then aligns
alleles by letter, every outcome beta ends up with the wrong sign while the
results look entirely plausible. To establish which convention a file uses:

1. **Read the README/header.** `ALLELE1`/`A1FREQ` (regenie) or
   `effect_allele` are explicit; `REF`/`ALT` means the effect is per ALT.
2. **Rule out a minor allele frequency.** A MAF never exceeds 0.5, so if the
   frequency column does, it tracks one fixed allele:

   ```{r maf-check, eval = FALSE}
   frq <- data.table::fread(path, select = "FRQ")[[1]]
   mean(frq > 0.5)
   #> [1] 0.155   # not a MAF -- describes a fixed allele (A1 or A2?)
   ```
3. **Look up two or three variants.** Compare `FRQ` with a known population
   frequency (gnomAD, Ensembl, or `plink --freq` on your LD panel) to see
   whether it describes `A1` or `A2`, and check a variant whose effect
   direction is established for the trait. For AMD, CFH rs1061170: the C
   allele is the risk allele (OR about 2.5). In the IAMDGC file that row reads
   `A1 = C, A2 = T, BETA = -0.87` -- read per `A1` the best-replicated AMD
   risk allele would be protective, so `BETA` is per `A2`.

If `A1` is REF, swap the mapping (user aliases take precedence):

```{r a1-swap, eval = FALSE}
col_map <- list(pval = "PVALUE", effect_allele = "A2", other_allele = "A1")
```

### Step 4: call format_gwas()

```{r format-gwas, eval = FALSE}
outcome <- format_gwas(
  path         = path,
  phenotype_id = "AMD",
  col_map      = col_map
)

exposure <- format_gwas(
  path         = "genomics_data/exposure_GWAS/RPS/rps.regenie",
  phenotype_id = "RPS",
  type         = "exposure",
  col_map      = list(rsids = "ID", pval = "LOG10P"),
  log10_pval   = TRUE
)
```

### Step 5: sanity-check the output

```{r sanity, eval = FALSE}
outcome[outcome$rsids == "rs1061170", c("effect_allele", "other_allele", "beta", "eaf")]
#>   effect_allele other_allele  beta   eaf
#> 1             C            T  0.87 0.38   # risk allele now has a positive beta
range(outcome$eaf, na.rm = TRUE)
```

### Step 6: the harmonisation safety net

`run_mr()` and `run_coloc()` run an allele orientation check after
harmonisation. Among non-palindromic SNPs with both allele frequencies, it
counts how many have `eaf.exposure` closer to `1 - eaf.outcome` than to
`eaf.outcome`. Different ancestries produce scatter; swapped alleles produce
systematic complementarity, so a clear majority (over 70% of at least 10 SNPs)
fails the check. `run_mr()` harmonises its instruments together with up to
1000 further SNPs shared by the two datasets, so the check works even for a
cis-MR with three instruments.

```{r allele-check, eval = FALSE}
result <- run_mr(exposure, "RPS", outcome, "AMD", instruments = ivs)
#> Error in `run_mr()`:
#> ! Possible effect/other allele mis-assignment between "RPS" and "AMD":
#>   987/1003 (98%) non-palindromic SNPs have eaf.exposure closer to
#>   1 - eaf.outcome than to eaf.outcome.
#> i This usually means one of the two datasets labels A1/A2 as REF/ALT ...
#> i Worst offenders (effect allele in brackets):
#>   rs117756744 (G): eaf.exposure = 0.979, eaf.outcome = 0.015, 1 - eaf.outcome = 0.985
#>   ...
#> ! Palindromic SNPs in this pair were strand-aligned from these mis-assigned
#>   frequencies, so their alignment is unreliable even where the beta looks unchanged.
#> ! The check cannot tell WHICH dataset is wrong. ...
#> > Fix: re-run `format_gwas()` on the offending dataset with
#>   `col_map = list(effect_allele = "A2", other_allele = "A1")` ...

chk <- last_allele_check()          # full record, stored on pass and fail
chk$status
chk$variants[order(-chk$variants$score), ][1:10, ]
```

Two things the check cannot do. It cannot tell *which* dataset is wrong --
the comparison is symmetric -- so break the tie with an independent frequency
reference (`plink --freq` on the LD panel, or `ref_frq` in `format_gwas()`),
a known-direction variant as in Step 3, or provenance (a dataset that has
harmonised cleanly against other outcomes is not the suspect). And it does not
flag palindromic SNPs: their strand is resolved *from* the frequencies, so a
mis-assigned effect allele and a mis-assigned frequency cancel and the beta can
come out identical either way. Once the check fails, treat every palindromic
alignment in that pair as unreliable and re-run after fixing the file.

Use `allele_check = "warn"` to continue with a warning, or `"none"` to skip
the condition (the record is still stored). Both are visible in
`result$params$allele_check`, which keeps the decision reviewable.

### When the check cannot run

The check needs allele frequencies in *both* datasets. OR-only GWAS files are
common and often ship without a frequency column, so no verdict is possible.
That is **not** a pass, and it says so at warning level (issue #21):

```{r allele-unverified, eval = FALSE}
#> Warning:
#> ! Allele orientation could not be checked for "RPS" vs "Malignant melanoma":
#>   no harmonised variants carry allele frequencies in both datasets.
#> ! Orientation is unverified for this pair. A mis-assigned effect allele
#>   inverts every estimate and leaves no other trace, so an unchecked pair is
#>   not the same as a clean one.
#> i Verify it by anchoring on a variant whose effect direction for this trait
#>   is established beyond doubt -- no frequencies needed. See `?format_gwas`,
#>   section What does A1 mean? (subsection If the check cannot run).
```

Verify orientation by hand instead: anchor on a variant whose direction for
the trait is beyond doubt and see whether the file agrees with the literature.
For a cutaneous melanoma GWAS, rs1805007 (MC1R), rs16891982 (SLC45A2) and
rs12203592 (IRF4) settle it at once -- read with the alleles swapped, all
three would contradict well-replicated melanoma genetics. `?format_gwas`
(*If the check cannot run*) works the example through and covers the two
checks that come free in many designs: a positive control that comes out the
wrong way makes orientation a prime suspect, and two independent GWAS of the
same trait should agree in direction at shared variants.

Once confirmed that way, `allele_check = "none"` silences the warning for that
pair -- and, being recorded in `result$params$allele_check`, the decision stays
reviewable.

### Inspecting what harmonisation did

`summary()` prints a harmonisation breakdown, and the full unfiltered frame is
on the result object as `$harmonisation`:

```{r harmonisation-diagnostics, eval = FALSE}
summary(result)
#> -- Harmonisation --
#> * 120 candidate SNPs -> 95 kept, 25 dropped
#> * Flagged: 10 palindromic, 8 ambiguous, 7 incompatible alleles

# Which variants, and why
h <- result$harmonisation
h[!h$mr_keep, c("SNP", "palindromic", "ambiguous", "remove")]
```

The flags overlap -- every ambiguous variant is palindromic -- so they do not
sum to the dropped total, and which of them actually costs a variant its place
depends on `harmonise_action` (below). A variant can also be dropped for
missing beta/se, which no allele flag shows; `summary()` reports those as
*incomplete beta/se*.

For `run_mr()`, `$instruments` remains the kept set the estimates are computed
from, and `$harmonisation` explains the rest. For `run_coloc()`,
`$harmonised_data` is still exactly the SNPs the analysis ran on -- row-aligned
with the coloc datasets, which the plots depend on -- and `$harmonisation` sits
alongside it.

### Choosing how palindromes are harmonised

`harmonise_action` (on both `run_mr()` and `run_coloc()`) sets the `action`
level `TwoSampleMR::harmonise_data()` uses:

| `harmonise_action` | Behaviour |
|---|---|
| 1 | Assume all alleles are on the forward strand: no frequency-based flip |
| 2 (default) | Infer the positive strand, resolving palindromes from allele frequencies |
| 3 | As 2, but drop every palindromic, ambiguous or incompatible SNP |

Only palindrome handling differs -- non-palindromic variants are aligned by
allele letter at every level, so the orientation check's verdict is the same
whichever you pick.

Reach for `3` when the frequencies level `2` relies on cannot be trusted.
That is exactly the situation after a failed orientation check: a swapped
effect allele and a swapped frequency cancel for a palindromic SNP, so its
strand call is unreliable even where the beta looks unchanged. It also covers
the case above, where no frequencies are available at all -- level `2` has
nothing to resolve palindromes with, so dropping them is the honest choice:

```{r harmonise-action, eval = FALSE}
result <- run_mr(
  exposure, "RPS", outcome, "Malignant melanoma",
  instruments = ivs,
  harmonise_action = 3,   # drop palindromic SNPs rather than guess their strand
  allele_check = "none"   # orientation confirmed by anchor variant instead
)
result$params$harmonise_action
```

The cost is instruments: dropping palindromes can remove a meaningful share
of a small cis-MR's SNPs, so check `result$results$nsnp` afterwards.

## Running MR Analyses

### Cis-MR (quick start with API clumping)

A cis-MR needs an exposure, an outcome and the gene's region. Setting
`instrument_region` selects genome-wide significant variants within
100 kb of it, and with no `bfile` they are LD-clumped through the OpenGWAS
API:

```{r cis-mr-api, eval = FALSE}
# Not run: needs network access and an OpenGWAS token -- see
# ieugwasr::get_opengwas_jwt().
mr_api <- run_mr(
  exposure = cd40_exposure,
  exposure_id = "CD40",
  outcome = sjogren_outcome,
  outcome_id = "SjD",
  instrument_region = list(chromosome = "20", start = 44746911, end = 44758502)
)
```

Without a `methods` argument `run_mr()` runs its default set: random-effects
IVW, MR-Egger, weighted median, MR-PRESSO, contamination mixture and Steiger
filtering.

### Cis-MR with local LD reference (recommended)

With a local panel the same analysis is reproducible and offline. This is the
analysis the rest of the MR sections inspect:

```{r cis-mr, eval = has_ld}
set.seed(1) # the weighted median's standard error is bootstrapped
mr_res <- run_mr(
  exposure = cd40_exposure,
  exposure_id = "CD40",
  outcome = sjogren_outcome,
  outcome_id = "SjD",
  instrument_region = list(chromosome = "20", start = 44746911, end = 44758502),
  rsq_thresh = 0.3,
  bfile = bfile,
  methods = c(
    "ivw_random", "ivw_fixed", "egger", "weighted_median",
    "steiger", "pleiotropy", "heterogeneity", "loo"
  )
)
```

Two arguments are worth a note. `rsq_thresh = 0.3` clumps far more leniently
than the default of 0.001: against real LD the default keeps a single CD40
instrument (a Wald ratio, with no sensitivity analyses possible), while 0.3
keeps `r num(nrow(mr_res$instruments))`. Instruments that correlated are
counted as independent by standard IVW, which is what
[LD-corrected MR](#ld-corrected-mr) below is for. And `methods` names every
estimator and diagnostic explicitly, so the analysis says what it ran.

`$results` holds one row per estimator:

```{r cis-mr-results, eval = has_ld}
mr_res$results[, c("method", "nsnp", "b", "se", "pval", "or")]
```

### Sensitivity analyses

Every estimator makes a different assumption about invalid instruments, so
agreement between them is the reassurance, and disagreement the lead:

- **IVW** (`ivw_random`, `ivw_fixed`) assumes every instrument is valid, or
  that pleiotropy averages to zero.
- **MR-Egger** (`egger`) allows directional pleiotropy through an intercept,
  at the cost of much lower precision; `pleiotropy` tests that intercept
  against zero.
- **Weighted median** (`weighted_median`) is consistent if at least half of
  the weight comes from valid instruments.
- **MR-PRESSO** (`presso`) and the **contamination mixture** (`conmix`) detect
  or down-weight outlying instruments. Both are in the default `methods`;
  see `mr_methods()` below for everything available.

Here the IVW and weighted-median estimates are
`r num(mr_res$results$b[mr_res$results$method == "IVW (random effects)"])`
and
`r num(mr_res$results$b[mr_res$results$method == "Weighted median"])`;
Egger's slope, `r num(mr_res$results$b[mr_res$results$method == "MR Egger"])`,
is far less precise (SE
`r num(mr_res$results$se[mr_res$results$method == "MR Egger"])`). The Egger
intercept, Cochran's Q (`heterogeneity`, needs >= 2 instruments) and the
leave-one-out estimates (`loo`, needs >= 3) each have their own field:

```{r sensitivity, eval = has_ld}
mr_res$pleiotropy[, c("egger_intercept", "se", "pval")]
mr_res$heterogeneity[, c("method", "Q", "Q_df", "Q_pval")]
mr_res$loo[, c("SNP", "b", "se", "p")]
```

```{r loo-shift, include = FALSE, eval = has_ld}
loo <- mr_res$loo
loo_all <- loo$b[loo$SNP == "All"]
loo_snp <- loo[loo$SNP != "All", ]
loo_max <- loo_snp[which.max(abs(loo_snp$b - loo_all)), ]
```

The leave-one-out table re-fits IVW without each SNP in turn, and its last
row, `All`, is the full-set estimate. Here dropping
`r if (has_ld) loo_max$SNP else "[not computed]"` moves the estimate the
most, from `r num(loo_all)` to `r num(loo_max$b)` -- in a
small instrument set one variant can carry much of the answer.

### Which methods can I run?

`mr_methods()` returns the table `run_mr()` itself dispatches from -- the
shortcut names, what each one runs, whether it is a fixed- or random-effects
estimator, whether `ld_correct = TRUE` applies to it, how many instruments it
needs, and where its result lands on the `mr_result`:

```{r method-table}
tab <- mrpipeline::mr_methods()
# Backticks render the field names as code, which keeps MathJax from reading
# `$results` ... `$steiger` as an inline formula.
tab$output <- paste0("`", tab$output, "`")
knitr::kable(tab)
```

`mr_methods(detail = "full")` adds the automatic Wald-ratio path (used when
exactly one instrument survives), the raw TwoSampleMR passthrough (any
`TwoSampleMR::mr_method_list()$obj` name with no shortcut, e.g. `"mr_raps"`),
and the function that actually runs on each LD path. The same table is
rendered into `?run_mr`. Because `run_mr()` reads it rather than a separate
list, it cannot drift from what the function accepts.

The IVW shortcuts name the estimator explicitly: `ivw_random` is
multiplicative random effects (the standard error is inflated by
`max(RSE, 1)`, never deflated) and `ivw_fixed` is fixed effects (no
overdispersion adjustment). With under-dispersed instruments the two coincide
exactly. The former names `"ivw"` and `"ivw_fe"` still work but warn.

### LD-corrected MR

Cis-MR instruments are often in linkage disequilibrium with one another, and
standard IVW treats every instrument as an independent look at the causal
effect -- two correlated SNPs are counted twice and the standard error comes
out too small. `ld_correct = TRUE` computes the instruments' LD matrix from
`bfile`, re-orients it to the exposure's effect alleles, and fits the IVW and
Egger estimators by generalised least squares with that matrix as the weight
structure (through `MendelianRandomization` with `correl = TRUE`).

The example repeats the analysis above -- same region, same
`rsq_thresh = 0.3` -- with the correction switched on:

```{r ld-correct, eval = has_ld}
corrected <- run_mr(
  exposure = cd40_exposure,
  exposure_id = "CD40",
  outcome = sjogren_outcome,
  outcome_id = "SjD",
  instrument_region = list(chromosome = "20", start = 44746911, end = 44758502),
  rsq_thresh = 0.3,
  bfile = bfile,
  ld_correct = TRUE,
  methods = c("ivw_random", "ivw_fixed", "egger", "weighted_median", "heterogeneity")
)

corrected$results[, c("method", "b", "se", "model", "ld_corrected")]
```

```{r ld-numbers, include = FALSE, eval = has_ld}
est <- function(res, method, col = "b") res$results[[col]][res$results$method == method]
ivw_q <- function(res) {
  res$heterogeneity$Q[res$heterogeneity$method == "Inverse variance weighted"]
}

# Pairs whose ratio estimates are negatively correlated: sign(r_ij * bx_i * bx_j)
ld <- corrected$ld_matrix
bx <- corrected$instruments$beta.exposure[
  match(rownames(ld), corrected$instruments$SNP)
]
pair_sign <- sign(ld * outer(bx, bx))[upper.tri(ld)]
lead <- which(rownames(ld) == "rs4810485")
lead_sign <- sign(ld[lead, -lead] * bx[lead] * bx[-lead])
max_r <- max(abs(ld[upper.tri(ld)]))

# The prose below makes qualitative claims about these numbers; fail the build
# rather than publish them if the bundled data ever stop supporting them.
stopifnot(
  nrow(corrected$instruments) == nrow(mr_res$instruments),
  est(corrected, "IVW (random effects)", "se") < est(mr_res, "IVW (random effects)", "se"),
  est(corrected, "IVW (fixed effects)", "se") < est(mr_res, "IVW (fixed effects)", "se"),
  mean(pair_sign < 0) > 0.5,
  mean(lead_sign < 0) > 0.5,
  ivw_q(corrected) > ivw_q(mr_res)
)
```

Its instruments are the `r num(nrow(ld))` from before, with pairwise |r| up
to `r num(max_r, 2)` in the panel -- a typical lenient cis-MR set, and one
with something to correct for. Without `ld_correct` (`mr_res`, above) the
IVW estimate was `r num(est(mr_res, "IVW (random effects)"))` (random-effects
SE `r num(est(mr_res, "IVW (random effects)", "se"), 2)`, fixed-effects SE
`r num(est(mr_res, "IVW (fixed effects)", "se"), 2)`) and the Egger slope
`r num(est(mr_res, "MR Egger"))`: treating correlated instruments as
independent changes the answer, not just its precision.

Note that the corrected standard errors here are *smaller*. Correction does
not always widen them: what matters is not `r` itself but the correlation
between the SNPs' ratio estimates, which is `r` times the sign of
`beta.exposure_i * beta.exposure_j`. Where that is positive the two
instruments are counted twice over and correction widens the SE; where it is
negative their errors partly cancel and GLS extracts *more* information than
the independent-instruments model. In this set
`r num(sum(pair_sign < 0))` of the `r num(length(pair_sign))` pairs are
negative, including `r num(sum(lead_sign < 0))` of the
`r num(length(lead_sign))` involving rs4810485 -- the largest-effect, and so
most heavily weighted, instrument -- and the standard errors fall. Such a gain leans entirely on the
reference panel's LD matching the GWAS sample's, so treat a large drop with
caution.

Every `$results` row states the estimator that produced it: `model` says
whether it was a fixed- or random-effects fit, and `ld_corrected` whether the
LD matrix was used. Only `ivw_random`, `ivw_fixed` and `egger` have a
correlated form; any other estimator you request runs on the uncorrected
data and warns by name -- the warning above -- so `ld_correct` is never
silently ignored. `summary(corrected)` lists which methods it was and was
not applied to, and `print(corrected)` tags the primary row `[LD-corrected]`.

The diagnostics are corrected too. Cochran's Q, the Egger intercept and the
leave-one-out estimates are all answers to "how much do my instruments
disagree?", and that question changes once the instruments are correlated --
the generalised Q for a correlated set can differ materially from the
naive one. So under `ld_correct = TRUE`, `$heterogeneity`, `$pleiotropy`
and `$loo` come from the correlated fits, and each carries its own
`ld_corrected` column on both arms:

```{r ld-diagnostics, eval = has_ld}
corrected$heterogeneity[, c("method", "Q", "Q_df", "Q_pval", "ld_corrected")]
```

Uncorrected, the same instruments give an IVW Q of
`r num(ivw_q(mr_res), 4)` (p =
`r num(mr_res$heterogeneity$Q_pval[mr_res$heterogeneity$method == "Inverse variance weighted"], 2)`):
here the naive test understates the disagreement between correlated
instruments.

The `method` labels in `$heterogeneity` stay TwoSampleMR's on both arms
(Q does not depend on the fixed/random choice), so the column is the only
thing that differs. Steiger filtering is the one diagnostic left as-is: it
compares per-SNP r^2 values and involves no weight matrix.

Two things to know:

- **`ivw_random` and `ivw_fixed` mean the same thing at every instrument
  count.** `MendelianRandomization`'s own default would switch to fixed
  effects below 4 instruments -- a common situation in cis-MR -- so
  `run_mr()` pins the model explicitly on the shortcut. Ask for the estimator
  you want by name.
- **An LD-corrected run can have fewer instruments than the same call
  uncorrected.** Instruments absent from the reference panel, or with
  ambiguous palindromic alleles, are dropped when the matrix is aligned to
  the exposure's effect alleles.

Because of the second point, one `mr_result` is always one instrument set
under one weight matrix -- `ld_correct` is a single switch, not a request for
both arms. To compare corrected and uncorrected estimates, run both -- here
`mr_res` from above is the uncorrected arm -- and section them in one forest
plot:

```{r ld-compare, eval = has_gg, warning = FALSE, fig.width = 7, fig.height = 4}
forest_plot(list("Uncorrected" = mr_res, "LD-corrected" = corrected))
```

### Genome-wide MR

<!-- TODO: genome-wide example -->

### Excluding genomic regions

Use the `exclude_regions` argument to remove instruments falling in specific
genomic regions. This is commonly used to exclude the MHC region on chromosome
6, which can introduce spurious associations due to complex LD structure.

Supply a data frame with columns `chr`, `start`, and `end`. The bundled data
cover only the CD40 region on chromosome 20, so these two examples use
placeholder datasets and are not run:

```{r exclude-mhc, eval = FALSE}
# Exclude MHC region (GRCh37 coordinates: chr6:28,477,797-33,448,354)
mhc_grch37 <- data.frame(chr = "6", start = 28477797, end = 33448354)

result <- run_mr(
  exposure = exposure_data,
  exposure_id = "PCSK9",
  outcome = outcome_data,
  outcome_id = "CHD",
  exclude_regions = mhc_grch37
)

# GRCh38 coordinates: chr6:28,510,120-33,480,577
mhc_grch38 <- data.frame(chr = "6", start = 28510120, end = 33480577)
```

You can exclude multiple regions by stacking rows. For example, when studying
age-related macular degeneration (AMD), you might exclude both the CFH and
ARMS2/HTRA1 loci:

```{r exclude-multi, eval = FALSE}
amd_exclusions <- data.frame(
  chr = c("1", "10"),
  start = c(196621008, 124214077),
  end = c(196716634, 124274424)
)

result <- run_mr(
  exposure = exposure_data,
  exposure_id = "CFH",
  outcome = outcome_data,
  outcome_id = "AMD",
  exclude_regions = amd_exclusions
)
```

### Manual instrument sets

You can bypass automatic instrument selection by supplying a character vector
of rsIDs to the `instruments` argument -- for example, a pre-specified pair of
variants:

```{r manual-instruments, eval = has_ld}
manual_res <- run_mr(
  exposure = cd40_exposure,
  exposure_id = "CD40",
  outcome = sjogren_outcome,
  outcome_id = "SjD",
  instruments = c("rs4810485", "rs4813002"),
  methods = c("ivw_random", "egger", "weighted_median", "heterogeneity", "loo")
)
manual_res$results[, c("method", "nsnp", "b", "se")]
```

Manual instruments are used as given: no p-value threshold, no clumping, so
no `bfile` is needed. With only two of them, every method that needs three
or more instruments cannot run. Rather than failing, `run_mr()` records why in
`$methods_skipped` (see [Interpreting MR results](#interpreting-mr-results)).

By default, instruments missing from the exposure data produce a warning and
the analysis continues with the available SNPs:

```{r manual-missing, eval = has_ld}
manual_lax <- run_mr(
  exposure = cd40_exposure,
  exposure_id = "CD40",
  outcome = sjogren_outcome,
  outcome_id = "SjD",
  instruments = c("rs4810485", "rs4813002", "rs0000001"),
  methods = "ivw_random"
)
```

Set `instruments_strict = TRUE` to make that an error instead:

```{r manual-strict, eval = has_ld, error = TRUE}
run_mr(
  exposure = cd40_exposure,
  exposure_id = "CD40",
  outcome = sjogren_outcome,
  outcome_id = "SjD",
  instruments = c("rs4810485", "rs4813002", "rs0000001"),
  instruments_strict = TRUE
)
```

## Interpreting MR Results

Use `print()` for a one-line summary and `summary()` for full details:

```{r mr-inspect, eval = has_ld, message = TRUE}
mr_res
summary(mr_res)
```

Beyond the estimates, three things in that output decide how far to trust
them.

**Instrument strength (`$f_stats`).** Each instrument's F-statistic,
computed as `(beta.exposure / se.exposure)^2`, measures how strongly it
predicts the exposure. Weak instruments (conventionally F < 10) bias
two-sample MR towards the null -- or, where the exposure and outcome samples
overlap, towards the confounded observational association. cis-pQTLs are
usually very strong: here the weakest instrument has
F = `r num(mr_res$f_stats$min, 3)`.

**Steiger filtering (`$steiger`).** For each instrument, Steiger's test asks
whether it explains more variance in the exposure (`rsq.exposure`) than in
the outcome (`rsq.outcome`), as it should if it acts on the outcome *through*
the exposure. An instrument failing that (`steiger_dir = FALSE`) suggests
reverse causation or pleiotropy, and is a candidate for exclusion.
`summary()` reports how many pass; `$steiger` has the per-SNP detail:

```{r steiger, eval = has_ld}
mr_res$steiger[, c("SNP", "rsq.exposure", "rsq.outcome", "steiger_dir", "steiger_pval")]
```

Steiger needs the exposure's sample size; without it the test is skipped.

**Skipped methods (`$methods_skipped`).** A method that could not run is not
an error: `run_mr()` returns the rest and records a reason for each method it
skipped -- too few instruments, a missing sample size, a failed fit. Always
check it before reading the absence of a row as meaningful. For the
two-instrument manual run above:

```{r methods-skipped, eval = has_ld}
manual_res$methods_skipped
```

`mr_methods()` lists the minimum instrument count of every method.

### Plotting MR results

`plot()` produces diagnostic plots using TwoSampleMR plotting functions
(requires `ggplot2`): `"scatter"` (the default), `"forest"` (one row per
SNP), `"funnel"` and `"loo"` (leave-one-out). Each returns TwoSampleMR's
list of plots, one per exposure-outcome pair, so take `[[1]]` for a single
analysis:

```{r mr-plot, eval = has_gg, warning = FALSE, fig.width = 6, fig.height = 4.5}
# warning = FALSE: TwoSampleMR's leave-one-out plot draws an empty spacer
# row, which ggplot2 reports as a removed missing value
plot(mr_res, type = "scatter")[[1]]
plot(mr_res, type = "loo")[[1]]
```

`plot(mr_res, type = "forest")` shows one row per SNP. For a manuscript-style
forest plot summarising *methods* (IVW, MR-Egger, weighted median, ...) or
*outcomes* (a primary analysis alongside positive/negative controls), use
`forest_plot()` and `outcome_forest_plot()` instead.

### Forest plots across analyses

`forest_plot()` shows one row per method, for a single `mr_result` or for
several sectioned together. Fixed effects is listed above random effects by
default. Method labels do not encode LD correction (that lives in the
`ld_corrected` column), so LD-corrected results match the same defaults:

```{r forest-plot-single, eval = has_gg, warning = FALSE, fig.width = 7, fig.height = 3}
forest_plot(mr_res)
```

Pass a named list to section several results into one figure. The
[LD-corrected MR](#ld-corrected-mr) comparison above is one example; a
primary analysis alongside positive and negative controls is another. The
package bundles no control outcomes, so this and the next example use
placeholder results and are not run:

```{r forest-plot-sectioned, eval = FALSE}
forest_plot(list(
  Primary = mr_res,
  "Positive control" = positive_control_res,
  "Negative control" = negative_control_res
))
```

`outcome_forest_plot()` instead shows one row per *outcome*, for one or both
IVW model variants, optionally coloured and/or shaped by a grouping column
you attach yourself (e.g. which instrument was used). Build the combined
data frame from one or more results' `$results`, then plot:

```{r outcome-forest-plot, eval = FALSE}
combined <- dplyr::bind_rows(
  mr_res$results |> dplyr::mutate(subcategory = "Primary"),
  positive_control_res$results |> dplyr::mutate(subcategory = "Positive control"),
  negative_control_res$results |> dplyr::mutate(subcategory = "Negative control")
) |>
  dplyr::mutate(
    instrument = dplyr::if_else(exposure == "CD40", "GWS instrument", "Functional instrument")
  )

outcome_forest_plot(
  combined,
  xlab = "OR (95% CI)",
  method = c("IVW (random effects)", "IVW (fixed effects)"), # both models
  colour_by = "instrument",
  shape_by = "method",
  section_order = c("Primary", "Positive control", "Negative control")
)
```

`table_forest_plot()` shows one row per outcome with an inline result/CI/
p-value table alongside the plotted estimate, via the `forestplot` package
(requires `forestplot`) rather than `ggplot2`. It returns a `forestplot`
object -- render it yourself with `plot()`, so you control the output device
(e.g. wrap the call in `pdf()`/`dev.off()` to save a file). Its input is a
plain data frame, built here from made-up values to show the layout:

```{r table-forest-plot, eval = has_forestplot, fig.width = 8, fig.height = 2.5}
# Illustrative values, not results.
table_dat <- data.frame(
  exposure = "Genetically-proxied NLRP3 inhibition",
  outcome = c("Coronary heart disease", "Stroke", "Type 2 diabetes"),
  n_snps = c(8, 8, 8),
  estimate = c(0.92, 0.95, 1.03),
  lower = c(0.85, 0.88, 0.94),
  upper = c(0.99, 1.02, 1.13),
  p_value = c(0.02, 0.15, 0.5)
)

fp <- table_forest_plot(table_dat, null_value = 1, xlab = "OR (95% CI)")
plot(fp)
```

## Colocalization

MR asks whether the exposure affects the outcome; colocalisation asks
whether the two traits share a causal variant in the region at all. A cis-MR
estimate driven by a variant that merely sits in LD with a *different*
outcome variant is a classic false positive, and colocalisation is the check
for it.

### Quick colocalization (ABF only)

The simplest test uses Approximate Bayes Factors (ABF), which assume at most
one causal variant per trait. Supply the formatted exposure, the outcome in
`format_gwas()` outcome format, the gene region (padded by `coloc_window`,
10 kb by default) and an LD reference panel. Sample sizes are read from the
data when `exposure_n`/`outcome_n` are not given. Sjogren's disease is a
case-control outcome, so `outcome_type` and the case fraction `outcome_s`
are set too (see [Case-control outcomes](#case-control-outcomes)):

```{r coloc-abf, eval = has_ld, results = "hide"}
coloc_res <- run_coloc(
  exposure = cd40_exposure,
  exposure_id = "CD40",
  outcome = sjogren_outcome,
  outcome_id = "SjD",
  gene_chr = 20,
  gene_start = 44746911,
  gene_end = 44758502,
  outcome_type = "cc",
  outcome_s = 6098 / 41420,
  bfile = bfile,
  methods = "abf"
)
```

`coloc` itself prints its posteriors as it runs, so that chunk's printed
output is hidden here; its warning is kept, and is the first clue to the
result. Inspect the result object instead:

```{r coloc-abf-inspect, eval = has_ld, message = TRUE}
coloc_res
summary(coloc_res)

# Posterior probabilities directly
coloc_res$coloc_abf$summary
```

```{r coloc-abf-numbers, include = FALSE, eval = has_ld}
pp <- coloc_res$coloc_abf$summary
# coloc works from beta and its variance, so take the p-value the same way
h <- coloc_res$harmonised_data
min_p_out <- min(2 * pnorm(-abs(h$beta.outcome / h$se.outcome)))
# The interpretation below depends on H1 winning; fail the build otherwise.
stopifnot(names(which.max(pp[-1])) == "PP.H1.abf", min_p_out > 1e-4)
```

The five hypotheses are H0 (no association with either trait), H1 (exposure
only), H2 (outcome only), H3 (both, distinct causal variants) and H4 (both,
one shared variant). Here H1 dominates, PP.H1 =
`r num(pp[["PP.H1.abf"]], 2)` against PP.H4 =
`r num(pp[["PP.H4.abf"]], 2)`: CD40 has a strong cis-pQTL, but
Sjogren's disease has no signal in the window -- its smallest p-value there
is `r num(min_p_out, 2)`, which is the `minimum p value` coloc warned about.
So this is *absence of evidence* for colocalisation, not evidence against a
shared variant.

Note the ratio `PP.H4 / (PP.H3 + PP.H4)` that `print()` shows. It is close
to 1, but it only answers "*if* both traits had a signal, would it be
shared?" -- a question that does not arise when the outcome has none. Read it
only alongside PP.H3 + PP.H4 themselves.

### Full colocalization (SuSiE + signals)

ABF's single-causal-variant assumption breaks down in regions with several
independent signals. Adding `"susie"` and `"signals"` to `methods` fine-maps
each trait with SuSiE, then tests colocalisation across all pairs of credible
sets with `coloc.susie()`, and across conditionally independent signals with
`coloc.signals()`:

```{r coloc-full, eval = has_susie, results = "hide", warning = FALSE}
set.seed(1)
coloc_full <- run_coloc(
  exposure = cd40_exposure,
  exposure_id = "CD40",
  outcome = sjogren_outcome,
  outcome_id = "SjD",
  gene_chr = 20,
  gene_start = 44746911,
  gene_end = 44758502,
  outcome_type = "cc",
  outcome_s = 6098 / 41420,
  bfile = bfile,
  methods = c("abf", "susie", "signals")
)
```

```{r coloc-full-inspect, eval = has_susie, message = TRUE}
summary(coloc_full)
```

```{r coloc-full-numbers, include = FALSE, eval = has_susie}
sig_pp4 <- max(coloc_full$coloc_signals$summary$PP.H4.abf)
stopifnot(coloc_full$params$susie_ncs_out == 0L, sig_pp4 < 0.8)
```

With no outcome signal SuSiE finds `r if (has_susie) coloc_full$params$susie_ncs_exp else "[not computed]"`
credible set(s) for CD40 and none for Sjogren's disease, so there is nothing
for `coloc.susie()` to pair up. `run_coloc()` does not fail: it records the
reason in `$methods_skipped`, which `summary()` reports under *Skipped
methods*, and still runs `coloc.signals()`. As with ABF, check
`$methods_skipped` before reading an empty field as a negative result.

`coloc.signals()` puts its best pair at PP.H4 =
`r if (has_susie) num(sig_pp4, 2) else "[not computed]"`, higher than ABF's
but still short of the 0.8 usually asked of evidence for colocalisation --
and it rests on an outcome whose strongest variant in the window has
p = `r num(min_p_out, 2)`. The two methods agree that these data cannot
settle whether CD40 and Sjogren's disease share a variant here.

### Case-control outcomes

For a case-control trait, `outcome_type = "cc"` with `outcome_s`, the
proportion of cases (here 6,098 of 41,420 participants), tells `coloc` how
to scale the effect sizes; the default `"quant"` treats the outcome as a
continuous trait with standard deviation `outcome_sdY`. The same applies to
the exposure through `exposure_type` and `exposure_s`. Getting it wrong
changes the prior on effect sizes and so the posteriors:

```{r coloc-cc, eval = has_ld, results = "hide", warning = FALSE}
coloc_quant <- run_coloc(
  exposure = cd40_exposure,
  exposure_id = "CD40",
  outcome = sjogren_outcome,
  outcome_id = "SjD",
  gene_chr = 20,
  gene_start = 44746911,
  gene_end = 44758502,
  bfile = bfile,
  methods = "abf"
)
```

```{r coloc-cc-compare, eval = has_ld}
rbind(
  case_control = coloc_res$coloc_abf$summary,
  quantitative = coloc_quant$coloc_abf$summary
)
```

### Plotting coloc results

`plot()` for `coloc_result` objects supports three plot types:

```{r coloc-plot, eval = has_gg, fig.width = 7, fig.height = 4}
plot(coloc_res, type = "pp_bar")   # ABF posterior probabilities (default; requires ggplot2)
plot(coloc_res, type = "regional") # regional association plots (requires ggplot2)
```

`type = "locuszoom"` renders LD-coloured regional plots via the
`locuszoomr` package (requires `locuszoomr` and `ensembldb`, plus an
Ensembl annotation package matching your data's genome build, e.g.
`EnsDb.Hsapiens.v75` for GRCh37 -- **attached** with `library()`, not just
installed, since `locuszoomr::locus()` looks up a character `ens_db` on the
search path). It needs a local LD reference panel (`bfile`) and an
`ens_db` -- there is no genome-build field stored on `coloc_result`, so
`ens_db` must be supplied explicitly and must match whatever build `bfile`
and the coloc data actually use. Unlike the other two plot types, this one
draws directly to the current graphics device (like a base R plot) and
returns `NULL` -- open a device first if you want to save it. (Not run here:
it needs the Ensembl annotation package and writes a file.)

```{r coloc-plot-locuszoom, eval = FALSE}
library(EnsDb.Hsapiens.v75)

pdf("locuszoom_plot.pdf", width = 9, height = 8)
plot(
  coloc_res,
  type = "locuszoom",
  ens_db = "EnsDb.Hsapiens.v75",
  bfile = "/path/to/ld_reference"
)
dev.off()
```

## Two-Stage Batch Workflow

<!-- TODO: Stage 1 (IVW screen, all proteins) -> Stage 2 (sensitivity + coloc, p < 0.05) -->
<!-- TODO: reference run_decode_iop_pipeline.R pattern -->
