---
title: "mrpipeline User Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mrpipeline User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

<!-- TODO: brief package description and when to use it -->

## Gene Coordinate Lookup

Use `get_gene_coords()` to programmatically retrieve gene coordinates from
Ensembl via biomaRt. This is useful for defining cis regions in `run_mr()` and
`run_coloc()` without hard-coding coordinates.

```{r gene-coords, eval = FALSE}
# GRCh38 (default)
coords <- get_gene_coords(c("CD40", "APOE"))
coords
#> # A tibble: 2 × 4
#>   hgnc_symbol chromosome     start       end
#>   <chr>       <chr>          <int>     <int>
#> 1 APOE        19          44905791  44909393
#> 2 CD40        20          44746911  44758502

# GRCh37
coords_37 <- get_gene_coords("CD40", build = "grch37")
```

The returned tibble can feed directly into `run_mr()` and `run_coloc()`:

```{r use-coords, eval = FALSE}
cd40 <- get_gene_coords("CD40", build = "grch38")

mr_res <- run_mr(
  exposure = exposure_data,
  exposure_id = "CD40",
  outcome = outcome_data,
  outcome_id = "SjD",
  instrument_region = list(
    chromosome = cd40$chromosome,
    start = cd40$start,
    end = cd40$end
  )
)
```

## Controlling PLINK Resource Usage

When running many parallel jobs (e.g. on an HPC cluster), PLINK's default
behaviour of auto-detecting available threads and memory can cause problems —
each worker may try to reserve half the node's RAM. Use `plink_threads` and
`plink_memory` to cap resource usage per call.

You can set these per-call, via R options, or via environment variables:

```{r plink-resources, eval = FALSE}
# Per-call (highest priority)
result <- run_mr(
 exposure = exposure_data, exposure_id = "CD40",
 outcome = outcome_data, outcome_id = "SjD",
 bfile = "/path/to/ld_ref",
 plink_threads = 1,
 plink_memory = 2000
)

# Via R options (apply to all subsequent calls in the session)
options(
 mrpipeline.plink_threads = 1,
 mrpipeline.plink_memory = 2000
)

# Via environment variables (e.g. in .Renviron or a job script)
# MRPIPELINE_PLINK_THREADS=1
# MRPIPELINE_PLINK_MEMORY=2000
```

Priority order: explicit argument > R option > environment variable > PLINK
auto-detect (`NULL`). The same parameters are available in `run_coloc()`.

## Setting Up an LD Reference Panel

<!-- TODO: explain when a local bfile is needed (batch, ld_correct, coloc) vs API -->
<!-- TODO: download instructions for 1000 Genomes EUR panel -->
<!-- TODO: expected file structure (bfile prefix: .bed/.bim/.fam) -->

## Formatting Exposure Data

<!-- TODO: format_pqtl_decode() example -->
<!-- TODO: format_pqtl_ukbppp() example -->
<!-- TODO: format_single_cell_onek1k() example -->
<!-- TODO: TwoSampleMR::format_data() for other sources -->
<!-- TODO: required outcome columns: rsids, chr, pos, beta, se, eaf, pval, n, effect_allele, other_allele -->

## Running MR Analyses

### Cis-MR (quick start with API clumping)

<!-- TODO: minimal example without bfile -->

### Cis-MR with local LD reference (recommended)

<!-- TODO: full example with bfile -->

### Sensitivity analyses

<!-- TODO: Egger, weighted median, PRESSO, ConMix, Steiger, heterogeneity, leave-one-out examples -->

Add `"heterogeneity"` and/or `"loo"` to `methods` to get Cochran's Q
(`TwoSampleMR::mr_heterogeneity()`) and leave-one-out
(`TwoSampleMR::mr_leaveoneout()`) alongside the effect estimates:

```r
result <- run_mr(
  exposure = cd40_exposure,
  exposure_id = "CD40",
  outcome = sjogren_outcome,
  outcome_id = "SjD",
  instrument_region = list(chromosome = "20", start = 44746911, end = 44758502),
  bfile = bfile,
  methods = c("ivw", "egger", "weighted_median", "heterogeneity", "loo")
)

result$heterogeneity  # Q, Q_df, Q_pval per method (needs >= 2 instruments)
result$loo            # per-SNP + pooled "All" estimates (needs >= 3 instruments)
```

### LD-corrected IVW

<!-- TODO: ld_correct = TRUE example with bfile -->

### Genome-wide MR

<!-- TODO: genome-wide example -->

### Excluding genomic regions

Use the `exclude_regions` argument to remove instruments falling in specific
genomic regions. This is commonly used to exclude the MHC region on chromosome
6, which can introduce spurious associations due to complex LD structure.

Supply a data frame with columns `chr`, `start`, and `end`:

```{r exclude-mhc, eval = FALSE}
# Exclude MHC region (GRCh37 coordinates: chr6:28,477,797–33,448,354)
mhc_grch37 <- data.frame(chr = "6", start = 28477797, end = 33448354)

result <- run_mr(
  exposure = exposure_data,
  exposure_id = "PCSK9",
  outcome = outcome_data,
  outcome_id = "CHD",
  exclude_regions = mhc_grch37
)

# GRCh38 coordinates: chr6:28,510,120–33,480,577
mhc_grch38 <- data.frame(chr = "6", start = 28510120, end = 33480577)
```

You can exclude multiple regions by stacking rows. For example, when studying
age-related macular degeneration (AMD), you might exclude both the CFH and
ARMS2/HTRA1 loci:

```{r exclude-multi, eval = FALSE}
amd_exclusions <- data.frame(
  chr = c("1", "10"),
  start = c(196621008, 124214077),
  end = c(196716634, 124274424)
)

result <- run_mr(
  exposure = exposure_data,
  exposure_id = "CFH",
  outcome = outcome_data,
  outcome_id = "AMD",
  exclude_regions = amd_exclusions
)
```

### Manual instrument sets

You can bypass automatic instrument selection by supplying a character vector
of rsIDs to the `instruments` argument:

```{r manual-instruments, eval = FALSE}
result <- run_mr(
  exposure = exposure_data,
  exposure_id = "PCSK9",
  outcome = outcome_data,
  outcome_id = "CHD",
  instruments = c("rs11591147", "rs2479409", "rs11583680")
)
```

By default, instruments missing from the exposure data produce a warning but
the analysis continues with the available SNPs. Set `instruments_strict = TRUE`
to error instead:

```{r manual-strict, eval = FALSE}
result <- run_mr(
  exposure = exposure_data,
  exposure_id = "PCSK9",
  outcome = outcome_data,
  outcome_id = "CHD",
  instruments = c("rs11591147", "rs2479409"),
  instruments_strict = TRUE
)
```

## Interpreting MR Results

<!-- TODO: F-statistics and what they mean -->
<!-- TODO: Steiger filtering -->
<!-- TODO: methods_skipped -->

Use `print()` for a one-line summary and `summary()` for full details:

```{r mr-inspect, eval = FALSE}
mr_res
summary(mr_res)
```

### Plotting MR results

`plot()` produces diagnostic plots using TwoSampleMR plotting functions
(requires `ggplot2`):

```{r mr-plot, eval = FALSE}
plot(mr_res, type = "scatter") # scatter plot (default)
plot(mr_res, type = "forest")  # per-SNP forest plot
plot(mr_res, type = "funnel")  # funnel plot
```

`plot(mr_res, type = "forest")` shows one row per SNP. For a manuscript-style
forest plot summarising *methods* (IVW, MR-Egger, weighted median, ...) or
*outcomes* (a primary analysis alongside positive/negative controls), use
`forest_plot()` and `outcome_forest_plot()` instead.

### Forest plots across analyses

`forest_plot()` shows one row per method, for a single `mr_result` or for
several sectioned together. Fixed effects is listed above random effects by
default, and `"Inverse variance weighted"` is relabelled to `"IVW (random
effects)"` for display:

```{r forest-plot-single, eval = FALSE}
forest_plot(mr_res)
```

Pass a named list to section several results into one figure -- e.g. a
primary analysis alongside positive and negative controls:

```{r forest-plot-sectioned, eval = FALSE}
forest_plot(list(
  Primary = mr_res,
  "Positive control" = positive_control_res,
  "Negative control" = negative_control_res
))
```

`outcome_forest_plot()` instead shows one row per *outcome*, for one or both
IVW model variants, optionally coloured and/or shaped by a grouping column
you attach yourself (e.g. which instrument was used). Build the combined
data frame from one or more results' `$results`, then plot:

```{r outcome-forest-plot, eval = FALSE}
combined <- dplyr::bind_rows(
  mr_res$results |> dplyr::mutate(subcategory = "Primary"),
  positive_control_res$results |> dplyr::mutate(subcategory = "Positive control"),
  negative_control_res$results |> dplyr::mutate(subcategory = "Negative control")
) |>
  dplyr::mutate(
    instrument = dplyr::if_else(exposure == "CD40", "GWS instrument", "Functional instrument")
  )

outcome_forest_plot(
  combined,
  xlab = "OR (95% CI)",
  method = c("Inverse variance weighted", "IVW (fixed effects)"), # both models
  colour_by = "instrument",
  shape_by = "method",
  section_order = c("Primary", "Positive control", "Negative control")
)
```

## Colocalization

### Quick colocalization (ABF only)

The simplest colocalization test uses Approximate Bayes Factors (ABF). Supply
pre-formatted exposure data, raw outcome data, the gene region, and an LD
reference panel:

```{r coloc-abf, eval = FALSE}
result <- run_coloc(
  exposure = formatted_exposure,
  outcome = outcome_data,
  gene_chr = 20,
  gene_start = 44746911,
  gene_end = 44758502,
  exposure_n = 35559,
  outcome_n = 400000,
  bfile = "/path/to/ld_reference",
  methods = "abf"
)
print(result)
summary(result)

# Access posterior probabilities directly
result$coloc_abf$summary
```

### Full colocalization (SuSiE + signals)

For multi-signal colocalization, include `"susie"` and `"signals"` in the
`methods` argument. SuSiE fine-maps each trait independently, then
`coloc.susie()` and `coloc.signals()` test colocalization across all pairs
of credible sets:

```{r coloc-full, eval = FALSE}
result <- run_coloc(
  exposure = formatted_exposure,
  outcome = outcome_data,
  gene_chr = 20,
  gene_start = 44746911,
  gene_end = 44758502,
  exposure_n = 35559,
  outcome_n = 400000,
  bfile = "/path/to/ld_reference",
  methods = c("abf", "susie", "signals")
)
summary(result)
```

### Case-control outcomes

For case-control outcomes, set `outcome_type = "cc"` and provide the
proportion of cases via `outcome_s`:

```{r coloc-cc, eval = FALSE}
result <- run_coloc(
  exposure = formatted_exposure,
  outcome = outcome_data,
  gene_chr = 20,
  gene_start = 44746911,
  gene_end = 44758502,
  exposure_n = 35559,
  outcome_n = 400000,
  outcome_type = "cc",
  outcome_s = 0.3,
  bfile = "/path/to/ld_reference",
  methods = c("abf", "susie", "signals")
)
```

### Plotting coloc results

`plot()` for `coloc_result` objects supports two plot types (requires
`ggplot2`):

```{r coloc-plot, eval = FALSE}
plot(coloc_res, type = "pp_bar")    # bar chart of ABF posterior probabilities (default)
plot(coloc_res, type = "regional")  # regional association plots
```

## Two-Stage Batch Workflow

<!-- TODO: Stage 1 (IVW screen, all proteins) → Stage 2 (sensitivity + coloc, p < 0.05) -->
<!-- TODO: reference run_decode_iop_pipeline.R pattern -->
