---
title: "mrpipeline User Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mrpipeline User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

<!-- TODO: brief package description and when to use it -->

## Gene Coordinate Lookup

Use `get_gene_coords()` to programmatically retrieve gene coordinates from
Ensembl via biomaRt. This is useful for defining cis regions in `run_mr()` and
`run_coloc()` without hard-coding coordinates.

```{r gene-coords, eval = FALSE}
# GRCh38 (default)
coords <- get_gene_coords(c("CD40", "APOE"))
coords
#> # A tibble: 2 × 4
#>   hgnc_symbol chromosome     start       end
#>   <chr>       <chr>          <int>     <int>
#> 1 APOE        19          44905791  44909393
#> 2 CD40        20          44746911  44758502

# GRCh37
coords_37 <- get_gene_coords("CD40", build = "grch37")
```

The returned tibble can feed directly into `run_mr()` and `run_coloc()`:

```{r use-coords, eval = FALSE}
cd40 <- get_gene_coords("CD40", build = "grch38")

mr_res <- run_mr(
  exposure = exposure_data,
  exposure_id = "CD40",
  outcome = outcome_data,
  outcome_id = "SjD",
  instrument_region = list(
    chromosome = cd40$chromosome,
    start = cd40$start,
    end = cd40$end
  )
)
```

## Controlling PLINK Resource Usage

When running many parallel jobs (e.g. on an HPC cluster), PLINK's default
behaviour of auto-detecting available threads and memory can cause problems —
each worker may try to reserve half the node's RAM. Use `plink_threads` and
`plink_memory` to cap resource usage per call.

You can set these per-call, via R options, or via environment variables:

```{r plink-resources, eval = FALSE}
# Per-call (highest priority)
result <- run_mr(
  exposure = exposure_data, exposure_id = "CD40",
  outcome = outcome_data, outcome_id = "SjD",
  bfile = "/path/to/ld_ref",
  plink_threads = 1,
  plink_memory = 2000
)

# Via R options (apply to all subsequent calls in the session)
options(
  mrpipeline.plink_threads = 1,
  mrpipeline.plink_memory = 2000
)

# Via environment variables (e.g. in .Renviron or a job script) -- set
# MRPIPELINE_PLINK_THREADS to 1 and MRPIPELINE_PLINK_MEMORY to 2000
```

Priority order: explicit argument > R option > environment variable > PLINK
auto-detect (`NULL`). The same parameters are available in `run_coloc()`.

## Setting Up an LD Reference Panel

<!-- TODO: explain when a local bfile is needed (batch, ld_correct, coloc) vs API -->
<!-- TODO: download instructions for 1000 Genomes EUR panel -->
<!-- TODO: expected file structure (bfile prefix: .bed/.bim/.fam) -->

## Formatting GWAS Data

`run_mr()` and `run_coloc()` take an **exposure** in TwoSampleMR format
(`SNP`, `beta.exposure`, `se.exposure`, `effect_allele.exposure`,
`other_allele.exposure`, `eaf.exposure`, `pval.exposure`, ...) and an
**outcome** in the `format_gwas()` outcome schema (`rsids`, `chr`, `pos`,
`beta`, `se`, `eaf`, `pval`, `n`, `effect_allele`, `other_allele`,
`phenotype`). `format_gwas()` produces either from almost any summary
statistics file. Source-specific wrappers exist for deCODE
(`format_pqtl_decode()`), UKB-PPP (`format_pqtl_ukbppp()`) and OneK1K
single-cell eQTLs (`format_single_cell_onek1k()`); they already know their
source's allele convention, so the steps below are for everything else.

### Step 1: inspect the header

Look at the column names before formatting anything:

```{r inspect-header, eval = FALSE}
path <- "genomics_data/outcome_GWAS/AMD/IAMDGC_AMD.tsv.gz"
names(data.table::fread(path, nrows = 5))
#> [1] "SNP" "CHR" "POS" "A1" "A2" "BETA" "SE" "PVALUE" "N" "FRQ" ...
```

Compare them against the alias table in `?format_gwas` (section *Column
normalisation*). Columns in the table are renamed automatically.

### Step 2: map any columns the alias table does not know

Anything not recognised needs a `col_map` entry. Other switches:
`log10_pval = TRUE` for `-log10(p)` columns, `n = ...` when the file has no
sample-size column, `bim_path` when rsIDs (or chr/pos) are missing, and
`marker_col` for compound `CHR:POS:...` identifiers.

```{r col-map, eval = FALSE}
col_map <- list(pval = "PVALUE")
```

### Step 3: decide what A1 means

This is the step that is easy to skip and expensive to get wrong. Two
conventions share the names `A1`/`A2`:

| Convention | `A1` | `A2` | `BETA` / `FRQ` refer to |
|---|---|---|---|
| PLINK, regenie, METAL, most GWAS Catalog deposits | effect allele | other allele | `A1` |
| EPACTS, RAREMETAL, VCF-derived tables | REF | ALT | `A2` |

`format_gwas()` assumes the first. A file using the second is silently read
with effect and other allele swapped, and because harmonisation then aligns
alleles by letter, every outcome beta ends up with the wrong sign while the
results look entirely plausible. To establish which convention a file uses:

1. **Read the README/header.** `ALLELE1`/`A1FREQ` (regenie) or
   `effect_allele` are explicit; `REF`/`ALT` means the effect is per ALT.
2. **Rule out a minor allele frequency.** A MAF never exceeds 0.5, so if the
   frequency column does, it tracks one fixed allele:

   ```{r maf-check, eval = FALSE}
   frq <- data.table::fread(path, select = "FRQ")[[1]]
   mean(frq > 0.5)
   #> [1] 0.155   # not a MAF -- describes a fixed allele (A1 or A2?)
   ```
3. **Look up two or three variants.** Compare `FRQ` with a known population
   frequency (gnomAD, Ensembl, or `plink --freq` on your LD panel) to see
   whether it describes `A1` or `A2`, and check a variant whose effect
   direction is established for the trait. For AMD, CFH rs1061170: the C
   allele is the risk allele (OR about 2.5). In the IAMDGC file that row reads
   `A1 = C, A2 = T, BETA = -0.87` -- read per `A1` the best-replicated AMD
   risk allele would be protective, so `BETA` is per `A2`.

If `A1` is REF, swap the mapping (user aliases take precedence):

```{r a1-swap, eval = FALSE}
col_map <- list(pval = "PVALUE", effect_allele = "A2", other_allele = "A1")
```

### Step 4: call format_gwas()

```{r format-gwas, eval = FALSE}
outcome <- format_gwas(
  path         = path,
  phenotype_id = "AMD",
  col_map      = col_map
)

exposure <- format_gwas(
  path         = "genomics_data/exposure_GWAS/RPS/rps.regenie",
  phenotype_id = "RPS",
  type         = "exposure",
  col_map      = list(rsids = "ID", pval = "LOG10P"),
  log10_pval   = TRUE
)
```

### Step 5: sanity-check the output

```{r sanity, eval = FALSE}
outcome[outcome$rsids == "rs1061170", c("effect_allele", "other_allele", "beta", "eaf")]
#>   effect_allele other_allele  beta   eaf
#> 1             C            T  0.87 0.38   # risk allele now has a positive beta
range(outcome$eaf, na.rm = TRUE)
```

### Step 6: the harmonisation safety net

`run_mr()` and `run_coloc()` run an allele orientation check after
harmonisation. Among non-palindromic SNPs with both allele frequencies, it
counts how many have `eaf.exposure` closer to `1 - eaf.outcome` than to
`eaf.outcome`. Different ancestries produce scatter; swapped alleles produce
systematic complementarity, so a clear majority (over 70% of at least 10 SNPs)
fails the check. `run_mr()` harmonises its instruments together with up to
1000 further SNPs shared by the two datasets, so the check works even for a
cis-MR with three instruments.

```{r allele-check, eval = FALSE}
result <- run_mr(exposure, "RPS", outcome, "AMD", instruments = ivs)
#> Error in `run_mr()`:
#> ! Possible effect/other allele mis-assignment between "RPS" and "AMD":
#>   987/1003 (98%) non-palindromic SNPs have eaf.exposure closer to
#>   1 - eaf.outcome than to eaf.outcome.
#> i This usually means one of the two datasets labels A1/A2 as REF/ALT ...
#> i Worst offenders (effect allele in brackets):
#>   rs117756744 (G): eaf.exposure = 0.979, eaf.outcome = 0.015, 1 - eaf.outcome = 0.985
#>   ...
#> ! Palindromic SNPs in this pair were strand-aligned from these mis-assigned
#>   frequencies, so their alignment is unreliable even where the beta looks unchanged.
#> ! The check cannot tell WHICH dataset is wrong. ...
#> > Fix: re-run `format_gwas()` on the offending dataset with
#>   `col_map = list(effect_allele = "A2", other_allele = "A1")` ...

chk <- last_allele_check()          # full record, stored on pass and fail
chk$status
chk$variants[order(-chk$variants$score), ][1:10, ]
```

Two things the check cannot do. It cannot tell *which* dataset is wrong --
the comparison is symmetric -- so break the tie with an independent frequency
reference (`plink --freq` on the LD panel, or `ref_frq` in `format_gwas()`),
a known-direction variant as in Step 3, or provenance (a dataset that has
harmonised cleanly against other outcomes is not the suspect). And it does not
flag palindromic SNPs: their strand is resolved *from* the frequencies, so a
mis-assigned effect allele and a mis-assigned frequency cancel and the beta can
come out identical either way. Once the check fails, treat every palindromic
alignment in that pair as unreliable and re-run after fixing the file.

Use `allele_check = "warn"` to continue with a warning, or `"none"` to skip
the condition (the record is still stored). Both are visible in
`result$params$allele_check`, which keeps the decision reviewable.

### When the check cannot run

The check needs allele frequencies in *both* datasets. OR-only GWAS files are
common and often ship without a frequency column, so no verdict is possible.
That is **not** a pass, and it says so at warning level (issue #21):

```{r allele-unverified, eval = FALSE}
#> Warning:
#> ! Allele orientation could not be checked for "RPS" vs "Malignant melanoma":
#>   no harmonised variants carry allele frequencies in both datasets.
#> ! Orientation is unverified for this pair. A mis-assigned effect allele
#>   inverts every estimate and leaves no other trace, so an unchecked pair is
#>   not the same as a clean one.
#> i Verify it by anchoring on a variant whose effect direction for this trait
#>   is established beyond doubt -- no frequencies needed. See `?format_gwas`,
#>   section What does A1 mean? (subsection If the check cannot run).
```

Verify orientation by hand instead: anchor on a variant whose direction for
the trait is beyond doubt and see whether the file agrees with the literature.
For a cutaneous melanoma GWAS, rs1805007 (MC1R), rs16891982 (SLC45A2) and
rs12203592 (IRF4) settle it at once -- read with the alleles swapped, all
three would contradict well-replicated melanoma genetics. `?format_gwas`
(*If the check cannot run*) works the example through and covers the two
checks that come free in many designs: a positive control that comes out the
wrong way makes orientation a prime suspect, and two independent GWAS of the
same trait should agree in direction at shared variants.

Once confirmed that way, `allele_check = "none"` silences the warning for that
pair -- and, being recorded in `result$params$allele_check`, the decision stays
reviewable.

### Inspecting what harmonisation did

`summary()` prints a harmonisation breakdown, and the full unfiltered frame is
on the result object as `$harmonisation`:

```{r harmonisation-diagnostics, eval = FALSE}
summary(result)
#> -- Harmonisation --
#> * 120 candidate SNPs -> 95 kept, 25 dropped
#> * Flagged: 10 palindromic, 8 ambiguous, 7 incompatible alleles

# Which variants, and why
h <- result$harmonisation
h[!h$mr_keep, c("SNP", "palindromic", "ambiguous", "remove")]
```

The flags overlap -- every ambiguous variant is palindromic -- so they do not
sum to the dropped total, and which of them actually costs a variant its place
depends on `harmonise_action` (below). A variant can also be dropped for
missing beta/se, which no allele flag shows; `summary()` reports those as
*incomplete beta/se*.

For `run_mr()`, `$instruments` remains the kept set the estimates are computed
from, and `$harmonisation` explains the rest. For `run_coloc()`,
`$harmonised_data` is still exactly the SNPs the analysis ran on -- row-aligned
with the coloc datasets, which the plots depend on -- and `$harmonisation` sits
alongside it.

### Choosing how palindromes are harmonised

`harmonise_action` (on both `run_mr()` and `run_coloc()`) sets the `action`
level `TwoSampleMR::harmonise_data()` uses:

| `harmonise_action` | Behaviour |
|---|---|
| 1 | Assume all alleles are on the forward strand: no frequency-based flip |
| 2 (default) | Infer the positive strand, resolving palindromes from allele frequencies |
| 3 | As 2, but drop every palindromic, ambiguous or incompatible SNP |

Only palindrome handling differs -- non-palindromic variants are aligned by
allele letter at every level, so the orientation check's verdict is the same
whichever you pick.

Reach for `3` when the frequencies level `2` relies on cannot be trusted.
That is exactly the situation after a failed orientation check: a swapped
effect allele and a swapped frequency cancel for a palindromic SNP, so its
strand call is unreliable even where the beta looks unchanged. It also covers
the case above, where no frequencies are available at all -- level `2` has
nothing to resolve palindromes with, so dropping them is the honest choice:

```{r harmonise-action, eval = FALSE}
result <- run_mr(
  exposure, "RPS", outcome, "Malignant melanoma",
  instruments = ivs,
  harmonise_action = 3,   # drop palindromic SNPs rather than guess their strand
  allele_check = "none"   # orientation confirmed by anchor variant instead
)
result$params$harmonise_action
```

The cost is instruments: dropping palindromes can remove a meaningful share
of a small cis-MR's SNPs, so check `result$results$nsnp` afterwards.

## Running MR Analyses

### Cis-MR (quick start with API clumping)

<!-- TODO: minimal example without bfile -->

### Cis-MR with local LD reference (recommended)

<!-- TODO: full example with bfile -->

### Sensitivity analyses

<!-- TODO: Egger, weighted median, PRESSO, ConMix, Steiger, heterogeneity, leave-one-out examples -->

Add `"heterogeneity"` and/or `"loo"` to `methods` to get Cochran's Q
(`TwoSampleMR::mr_heterogeneity()`) and leave-one-out
(`TwoSampleMR::mr_leaveoneout()`) alongside the effect estimates:

```r
result <- run_mr(
  exposure = cd40_exposure,
  exposure_id = "CD40",
  outcome = sjogren_outcome,
  outcome_id = "SjD",
  instrument_region = list(chromosome = "20", start = 44746911, end = 44758502),
  rsq_thresh = 0.3,
  bfile = bfile,
  methods = c("ivw_random", "egger", "weighted_median", "heterogeneity", "loo")
)

result$heterogeneity  # Q, Q_df, Q_pval per method (needs >= 2 instruments)
result$loo            # per-SNP + pooled "All" estimates (needs >= 3 instruments)
```

### Which methods can I run?

`mr_methods()` returns the table `run_mr()` itself dispatches from -- the
shortcut names, what each one runs, whether it is a fixed- or random-effects
estimator, whether `ld_correct = TRUE` applies to it, how many instruments it
needs, and where its result lands on the `mr_result`:

```{r method-table}
tab <- mrpipeline::mr_methods()
# Backticks render the field names as code, which keeps MathJax from reading
# `$results` ... `$steiger` as an inline formula.
tab$output <- paste0("`", tab$output, "`")
knitr::kable(tab)
```

`mr_methods(detail = "full")` adds the automatic Wald-ratio path (used when
exactly one instrument survives), the raw TwoSampleMR passthrough (any
`TwoSampleMR::mr_method_list()$obj` name with no shortcut, e.g. `"mr_raps"`),
and the function that actually runs on each LD path. The same table is
rendered into `?run_mr`. Because `run_mr()` reads it rather than a separate
list, it cannot drift from what the function accepts.

The IVW shortcuts name the estimator explicitly: `ivw_random` is
multiplicative random effects (the standard error is inflated by
`max(RSE, 1)`, never deflated) and `ivw_fixed` is fixed effects (no
overdispersion adjustment). With under-dispersed instruments the two coincide
exactly. The former names `"ivw"` and `"ivw_fe"` still work but warn.

### LD-corrected MR

Cis-MR instruments are often in linkage disequilibrium with one another, and
standard IVW treats every instrument as an independent look at the causal
effect -- two correlated SNPs are counted twice and the standard error comes
out too small. `ld_correct = TRUE` computes the instruments' LD matrix from
`bfile`, re-orients it to the exposure's effect alleles, and fits the IVW and
Egger estimators by generalised least squares with that matrix as the weight
structure (through `MendelianRandomization` with `correl = TRUE`):

The example uses the LD reference panel bundled with the package (real 1000
Genomes EUR genotypes for the 50 CD40-region SNPs). Its instruments are
clumped at `rsq_thresh = 0.3` rather than the default 0.001: against real LD
the default leaves a single CD40 instrument, while 0.3 keeps seven with
pairwise |r| up to 0.53 -- a typical lenient cis-MR set, and one with
something to correct for.

```{r ld-correct, eval = FALSE}
bfile <- sub(
  "\\.bed$",
  "",
  system.file("extdata", "ld_ref.bed", package = "mrpipeline")
)

corrected <- run_mr(
  exposure = cd40_exposure,
  exposure_id = "CD40",
  outcome = sjogren_outcome,
  outcome_id = "SjD",
  instrument_region = list(chromosome = "20", start = 44746911, end = 44758502),
  rsq_thresh = 0.3,
  bfile = bfile,
  ld_correct = TRUE,
  methods = c("ivw_random", "ivw_fixed", "egger", "weighted_median", "heterogeneity")
)
#> Warning: "weighted_median" has no LD-corrected form; running uncorrected.

corrected$results[, c("method", "b", "se", "model", "ld_corrected")]
#>                 method     b     se  model ld_corrected
#> 1 IVW (random effects) 0.153 0.0580 random         TRUE
#> 2  IVW (fixed effects) 0.153 0.0273  fixed         TRUE
#> 3             MR Egger 0.455 0.2829 random         TRUE
#> 4      Weighted median 0.168 0.0693   <NA>        FALSE
```

The same call without `ld_correct` gives an IVW estimate of 0.104 (random-effects
SE 0.071, fixed-effects SE 0.050) and an Egger slope of 0.146: treating the
seven correlated instruments as independent changes the answer, not just its
precision.

Note that the corrected standard errors here are *smaller*. Correction does
not always widen them: what matters is not `r` itself but the correlation
between the SNPs' ratio estimates, which is `r` times the sign of
`beta.exposure_i * beta.exposure_j`. Where that is positive the two
instruments are counted twice over and correction widens the SE; where it is
negative their errors partly cancel and GLS extracts *more* information than
the independent-instruments model. In this set most pairs are negative -- the
largest-effect instrument, rs4810485, against almost all the others -- so the
standard errors fall. Such a gain leans entirely on the reference panel's LD
matching the GWAS sample's, so treat a large drop with caution.

Every `$results` row states the estimator that produced it: `model` says
whether it was a fixed- or random-effects fit, and `ld_corrected` whether the
LD matrix was used. Only `ivw_random`, `ivw_fixed` and `egger` have a
correlated form; any other estimator you request runs on the uncorrected
data and warns by name, so `ld_correct` is never silently ignored.
`summary(corrected)` lists which methods it was and was not applied to, and
`print(corrected)` tags the primary row `[LD-corrected]`.

The diagnostics are corrected too. Cochran's Q, the Egger intercept and the
leave-one-out estimates are all answers to "how much do my instruments
disagree?", and that question changes once the instruments are correlated --
the generalised Q for a correlated set can differ materially from the
naive one. So under `ld_correct = TRUE`, `$heterogeneity`, `$pleiotropy`
and `$loo` come from the correlated fits, and each carries its own
`ld_corrected` column on both arms:

```{r ld-diagnostics, eval = FALSE}
corrected$heterogeneity[, c("method", "Q", "Q_df", "Q_pval", "ld_corrected")]
#>                      method    Q Q_df   Q_pval ld_corrected
#> 1                  MR Egger 21.9    5 0.000552         TRUE
#> 2 Inverse variance weighted 27.1    6 0.000140         TRUE
```

Uncorrected, the same instruments give an IVW Q of 12.0 (p = 0.063): here the
naive test understates the disagreement between correlated instruments.

The `method` labels in `$heterogeneity` stay TwoSampleMR's on both arms
(Q does not depend on the fixed/random choice), so the column is the only
thing that differs. Steiger filtering is the one diagnostic left as-is: it
compares per-SNP r^2 values and involves no weight matrix.

Two things to know:

- **`ivw_random` and `ivw_fixed` mean the same thing at every instrument
  count.** `MendelianRandomization`'s own default would switch to fixed
  effects below 4 instruments -- a common situation in cis-MR -- so
  `run_mr()` pins the model explicitly on the shortcut. Ask for the estimator
  you want by name.
- **An LD-corrected run can have fewer instruments than the same call
  uncorrected.** Instruments absent from the reference panel, or with
  ambiguous palindromic alleles, are dropped when the matrix is aligned to
  the exposure's effect alleles.

Because of the second point, one `mr_result` is always one instrument set
under one weight matrix -- `ld_correct` is a single switch, not a request for
both arms. To compare corrected and uncorrected estimates, run both and
section them in one forest plot:

```{r ld-compare, eval = FALSE}
uncorrected <- run_mr(
  exposure = cd40_exposure,
  exposure_id = "CD40",
  outcome = sjogren_outcome,
  outcome_id = "SjD",
  instrument_region = list(chromosome = "20", start = 44746911, end = 44758502),
  rsq_thresh = 0.3,
  bfile = bfile,
  methods = c("ivw_random", "ivw_fixed", "egger", "weighted_median", "heterogeneity")
)

forest_plot(list("Uncorrected" = uncorrected, "LD-corrected" = corrected))
```

### Genome-wide MR

<!-- TODO: genome-wide example -->

### Excluding genomic regions

Use the `exclude_regions` argument to remove instruments falling in specific
genomic regions. This is commonly used to exclude the MHC region on chromosome
6, which can introduce spurious associations due to complex LD structure.

Supply a data frame with columns `chr`, `start`, and `end`:

```{r exclude-mhc, eval = FALSE}
# Exclude MHC region (GRCh37 coordinates: chr6:28,477,797–33,448,354)
mhc_grch37 <- data.frame(chr = "6", start = 28477797, end = 33448354)

result <- run_mr(
  exposure = exposure_data,
  exposure_id = "PCSK9",
  outcome = outcome_data,
  outcome_id = "CHD",
  exclude_regions = mhc_grch37
)

# GRCh38 coordinates: chr6:28,510,120–33,480,577
mhc_grch38 <- data.frame(chr = "6", start = 28510120, end = 33480577)
```

You can exclude multiple regions by stacking rows. For example, when studying
age-related macular degeneration (AMD), you might exclude both the CFH and
ARMS2/HTRA1 loci:

```{r exclude-multi, eval = FALSE}
amd_exclusions <- data.frame(
  chr = c("1", "10"),
  start = c(196621008, 124214077),
  end = c(196716634, 124274424)
)

result <- run_mr(
  exposure = exposure_data,
  exposure_id = "CFH",
  outcome = outcome_data,
  outcome_id = "AMD",
  exclude_regions = amd_exclusions
)
```

### Manual instrument sets

You can bypass automatic instrument selection by supplying a character vector
of rsIDs to the `instruments` argument:

```{r manual-instruments, eval = FALSE}
result <- run_mr(
  exposure = exposure_data,
  exposure_id = "PCSK9",
  outcome = outcome_data,
  outcome_id = "CHD",
  instruments = c("rs11591147", "rs2479409", "rs11583680")
)
```

By default, instruments missing from the exposure data produce a warning but
the analysis continues with the available SNPs. Set `instruments_strict = TRUE`
to error instead:

```{r manual-strict, eval = FALSE}
result <- run_mr(
  exposure = exposure_data,
  exposure_id = "PCSK9",
  outcome = outcome_data,
  outcome_id = "CHD",
  instruments = c("rs11591147", "rs2479409"),
  instruments_strict = TRUE
)
```

## Interpreting MR Results

<!-- TODO: F-statistics and what they mean -->
<!-- TODO: Steiger filtering -->
<!-- TODO: methods_skipped -->

Use `print()` for a one-line summary and `summary()` for full details:

```{r mr-inspect, eval = FALSE}
mr_res
summary(mr_res)
```

### Plotting MR results

`plot()` produces diagnostic plots using TwoSampleMR plotting functions
(requires `ggplot2`):

```{r mr-plot, eval = FALSE}
plot(mr_res, type = "scatter") # scatter plot (default)
plot(mr_res, type = "forest")  # per-SNP forest plot
plot(mr_res, type = "funnel")  # funnel plot
```

`plot(mr_res, type = "forest")` shows one row per SNP. For a manuscript-style
forest plot summarising *methods* (IVW, MR-Egger, weighted median, ...) or
*outcomes* (a primary analysis alongside positive/negative controls), use
`forest_plot()` and `outcome_forest_plot()` instead.

### Forest plots across analyses

`forest_plot()` shows one row per method, for a single `mr_result` or for
several sectioned together. Fixed effects is listed above random effects by
default. Method labels do not encode LD correction (that lives in the
`ld_corrected` column), so LD-corrected results match the same defaults:

```{r forest-plot-single, eval = FALSE}
forest_plot(mr_res)
```

Pass a named list to section several results into one figure -- e.g. a
primary analysis alongside positive and negative controls:

```{r forest-plot-sectioned, eval = FALSE}
forest_plot(list(
  Primary = mr_res,
  "Positive control" = positive_control_res,
  "Negative control" = negative_control_res
))
```

`outcome_forest_plot()` instead shows one row per *outcome*, for one or both
IVW model variants, optionally coloured and/or shaped by a grouping column
you attach yourself (e.g. which instrument was used). Build the combined
data frame from one or more results' `$results`, then plot:

```{r outcome-forest-plot, eval = FALSE}
combined <- dplyr::bind_rows(
  mr_res$results |> dplyr::mutate(subcategory = "Primary"),
  positive_control_res$results |> dplyr::mutate(subcategory = "Positive control"),
  negative_control_res$results |> dplyr::mutate(subcategory = "Negative control")
) |>
  dplyr::mutate(
    instrument = dplyr::if_else(exposure == "CD40", "GWS instrument", "Functional instrument")
  )

outcome_forest_plot(
  combined,
  xlab = "OR (95% CI)",
  method = c("IVW (random effects)", "IVW (fixed effects)"), # both models
  colour_by = "instrument",
  shape_by = "method",
  section_order = c("Primary", "Positive control", "Negative control")
)
```

`table_forest_plot()` shows one row per outcome with an inline result/CI/
p-value table alongside the plotted estimate, via the `forestplot` package
(requires `forestplot`) rather than `ggplot2`. It returns a `forestplot`
object -- render it yourself with `plot()`, so you control the output device
(e.g. wrap the call in `pdf()`/`dev.off()` to save a file):

```{r table-forest-plot, eval = FALSE}
table_dat <- data.frame(
  exposure = "Genetically-proxied NLRP3 inhibition",
  outcome = c("Coronary heart disease", "Stroke", "Type 2 diabetes"),
  n_snps = c(8, 8, 8),
  estimate = c(0.92, 0.95, 1.03),
  lower = c(0.85, 0.88, 0.94),
  upper = c(0.99, 1.02, 1.13),
  p_value = c(0.02, 0.15, 0.5)
)

fp <- table_forest_plot(table_dat, null_value = 1, xlab = "OR (95% CI)")
plot(fp)
```

## Colocalization

### Quick colocalization (ABF only)

The simplest colocalization test uses Approximate Bayes Factors (ABF). Supply
pre-formatted exposure data, raw outcome data, the gene region, and an LD
reference panel:

```{r coloc-abf, eval = FALSE}
result <- run_coloc(
  exposure = formatted_exposure,
  outcome = outcome_data,
  gene_chr = 20,
  gene_start = 44746911,
  gene_end = 44758502,
  exposure_n = 35559,
  outcome_n = 400000,
  bfile = "/path/to/ld_reference",
  methods = "abf"
)
print(result)
summary(result)

# Access posterior probabilities directly
result$coloc_abf$summary
```

### Full colocalization (SuSiE + signals)

For multi-signal colocalization, include `"susie"` and `"signals"` in the
`methods` argument. SuSiE fine-maps each trait independently, then
`coloc.susie()` and `coloc.signals()` test colocalization across all pairs
of credible sets:

```{r coloc-full, eval = FALSE}
result <- run_coloc(
  exposure = formatted_exposure,
  outcome = outcome_data,
  gene_chr = 20,
  gene_start = 44746911,
  gene_end = 44758502,
  exposure_n = 35559,
  outcome_n = 400000,
  bfile = "/path/to/ld_reference",
  methods = c("abf", "susie", "signals")
)
summary(result)
```

### Case-control outcomes

For case-control outcomes, set `outcome_type = "cc"` and provide the
proportion of cases via `outcome_s`:

```{r coloc-cc, eval = FALSE}
result <- run_coloc(
  exposure = formatted_exposure,
  outcome = outcome_data,
  gene_chr = 20,
  gene_start = 44746911,
  gene_end = 44758502,
  exposure_n = 35559,
  outcome_n = 400000,
  outcome_type = "cc",
  outcome_s = 0.3,
  bfile = "/path/to/ld_reference",
  methods = c("abf", "susie", "signals")
)
```

### Plotting coloc results

`plot()` for `coloc_result` objects supports three plot types:

```{r coloc-plot, eval = FALSE}
plot(coloc_res, type = "pp_bar")    # bar chart of ABF posterior probabilities (default; requires ggplot2)
plot(coloc_res, type = "regional")  # regional association plots (requires ggplot2)
```

`type = "locuszoom"` renders LD-coloured regional plots via the
`locuszoomr` package (requires `locuszoomr` and `ensembldb`, plus an
Ensembl annotation package matching your data's genome build, e.g.
`EnsDb.Hsapiens.v75` for GRCh37 -- **attached** with `library()`, not just
installed, since `locuszoomr::locus()` looks up a character `ens_db` on the
search path). It needs a local LD reference panel (`bfile`) and an
`ens_db` -- there is no genome-build field stored on `coloc_result`, so
`ens_db` must be supplied explicitly and must match whatever build `bfile`
and the coloc data actually use. Unlike the other two plot types, this one
draws directly to the current graphics device (like a base R plot) and
returns `NULL` -- open a device first if you want to save it:

```{r coloc-plot-locuszoom, eval = FALSE}
library(EnsDb.Hsapiens.v75)

pdf("locuszoom_plot.pdf", width = 9, height = 8)
plot(
  coloc_res,
  type = "locuszoom",
  ens_db = "EnsDb.Hsapiens.v75",
  bfile = "/path/to/ld_reference"
)
dev.off()
```

## Two-Stage Batch Workflow

<!-- TODO: Stage 1 (IVW screen, all proteins) → Stage 2 (sensitivity + coloc, p < 0.05) -->
<!-- TODO: reference run_decode_iop_pipeline.R pattern -->
