---
title: "Getting Started with mrpipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mrpipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# The examples need the bundled LD panel and a PLINK binary, which
# genetics.binaRies downloads on first use; without them the code is shown
# with its output omitted rather than failing the build.
has_ld <- local({
  bed <- system.file("extdata", "ld_ref.bed", package = "mrpipeline")
  plink <- tryCatch(genetics.binaRies::get_plink_binary(), error = function(e) "")
  nzchar(bed) && file.exists(plink)
})
has_gg <- has_ld && requireNamespace("ggplot2", quietly = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # run_mr()/run_coloc() narrate their progress through messages
  message = FALSE,
  eval = has_ld
)
```

mrpipeline provides a streamlined interface for Mendelian randomisation (MR)
and colocalization analysis, with a focus on proteomic GWAS data from deCODE
and UKB-PPP. It wraps TwoSampleMR, coloc, and MendelianRandomization into a
consistent workflow with S3 result objects and built-in sensitivity analyses.

## Installation

Install from GitHub:

```r
# install.packages("pak")
pak::pak("BZuckerman97/mrpipeline")
```

## Quick start

```{r, eval = TRUE}
library(mrpipeline)
```

### Mendelian randomisation

mrpipeline ships with bundled test datasets for CD40 protein and Sjogren's
disease. Use these to explore the package without any external data.

```{r}
# Bundled datasets: cd40_exposure (formatted exposure), sjogren_outcome (outcome)
bfile <- sub("\\.bed$", "", system.file("extdata", "ld_ref.bed", package = "mrpipeline"))

# Run cis-MR (the weighted median's SE is bootstrapped, hence the seed)
set.seed(1)
mr_res <- run_mr(
  exposure = cd40_exposure,
  exposure_id = "CD40",
  outcome = sjogren_outcome,

  outcome_id = "SjD",
  instrument_region = list(chromosome = "20", start = 44746911, end = 44758502),
  rsq_thresh = 0.3,
  bfile = bfile,
  methods = c("ivw_random", "egger", "weighted_median")
)
```

Inspect the results:

```{r, message = TRUE}
mr_res
summary(mr_res)
```

and plot them (requires ggplot2; TwoSampleMR's plots come as a list, one per
exposure-outcome pair):

```{r, eval = has_gg, warning = FALSE, fig.width = 6, fig.height = 4.5}
# warning = FALSE: TwoSampleMR's forest plot draws an empty spacer row, which
# ggplot2 reports as a removed missing value
plot(mr_res, type = "scatter")[[1]]
plot(mr_res, type = "forest")[[1]]
```

The `mr_result` object stores the full results table, harmonised instruments,
F-statistics, Steiger filtering output, and any skipped methods — accessible
via `mr_res$results`, `mr_res$instruments`, etc.

### Colocalization

```{r, results = "hide", warning = FALSE}
coloc_res <- run_coloc(
  exposure = cd40_exposure,
  exposure_id = "CD40",
  outcome = sjogren_outcome,
  outcome_id = "SjD",
  gene_chr = "20",
  gene_start = 44746911,
  gene_end = 44758502,
  bfile = bfile,
  methods = "abf"
)
```

```{r, message = TRUE}
coloc_res
summary(coloc_res)
```

```{r, eval = has_gg, fig.width = 7, fig.height = 4}
plot(coloc_res, type = "pp_bar")
```

The bundled outcome data are partly synthetic, so these numbers illustrate
the workflow rather than a finding. `vignette("mrpipeline-user-guide")` walks
through what each part of the output means.

### Gene coordinate lookup

Look up genomic coordinates for HGNC gene symbols via Ensembl (requires the
`biomaRt` package). Not run here, as it needs network access:

```{r, eval = FALSE}
coords <- get_gene_coords(c("CD40", "APOE"), build = "grch38")
coords
```

These coordinates can be passed directly to `run_mr()` and `run_coloc()` via
the `instrument_region` and `gene_*` arguments.

## Formatting exposure data

mrpipeline includes formatters for common proteomic GWAS sources:

- `format_pqtl_decode()` — deCODE genetics
- `format_pqtl_ukbppp()` — UKB-PPP (Olink)
- `format_single_cell_onek1k()` — OneK1K single-cell eQTL

Each returns data formatted for TwoSampleMR, ready to pass to `run_mr()`.

## Further reading

- `vignette("mrpipeline-user-guide")` — detailed usage examples
- `vignette("mrpipeline-developer-guide")` — architecture and internals
