---
title: "mrpipeline User Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mrpipeline User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

<!-- TODO: brief package description and when to use it -->

## Setting Up an LD Reference Panel

<!-- TODO: explain when a local bfile is needed (batch, ld_correct, coloc) vs API -->
<!-- TODO: download instructions for 1000 Genomes EUR panel -->
<!-- TODO: expected file structure (bfile prefix: .bed/.bim/.fam) -->

## Formatting Exposure Data

<!-- TODO: format_pqtl_decode() example -->
<!-- TODO: format_pqtl_ukbppp() example -->
<!-- TODO: format_single_cell_onek1k() example -->
<!-- TODO: TwoSampleMR::format_data() for other sources -->
<!-- TODO: required outcome columns: rsids, chr, pos, beta, se, eaf, pval, n, effect_allele, other_allele -->

## Running MR Analyses

### Cis-MR (quick start with API clumping)

<!-- TODO: minimal example without bfile -->

### Cis-MR with local LD reference (recommended)

<!-- TODO: full example with bfile -->

### Sensitivity analyses

<!-- TODO: Egger, weighted median, PRESSO, ConMix, Steiger examples -->

### LD-corrected IVW

<!-- TODO: ld_correct = TRUE example with bfile -->

### Genome-wide MR

<!-- TODO: genome-wide example -->

### Excluding genomic regions

Use the `exclude_regions` argument to remove instruments falling in specific
genomic regions. This is commonly used to exclude the MHC region on chromosome
6, which can introduce spurious associations due to complex LD structure.

Supply a data frame with columns `chr`, `start`, and `end`:

```{r exclude-mhc, eval = FALSE}
# Exclude MHC region (GRCh37 coordinates: chr6:28,477,797–33,448,354)
mhc_grch37 <- data.frame(chr = "6", start = 28477797, end = 33448354)

result <- run_mr(
  exposure = exposure_data,
  exposure_id = "PCSK9",
  outcome = outcome_data,
  outcome_id = "CHD",
  exclude_regions = mhc_grch37
)

# GRCh38 coordinates: chr6:28,510,120–33,480,577
mhc_grch38 <- data.frame(chr = "6", start = 28510120, end = 33480577)
```

You can exclude multiple regions by stacking rows. For example, when studying
age-related macular degeneration (AMD), you might exclude both the CFH and
ARMS2/HTRA1 loci:

```{r exclude-multi, eval = FALSE}
amd_exclusions <- data.frame(
  chr = c("1", "10"),
  start = c(196621008, 124214077),
  end = c(196716634, 124274424)
)

result <- run_mr(
  exposure = exposure_data,
  exposure_id = "CFH",
  outcome = outcome_data,
  outcome_id = "AMD",
  exclude_regions = amd_exclusions
)
```

### Manual instrument sets

You can bypass automatic instrument selection by supplying a character vector
of rsIDs to the `instruments` argument:

```{r manual-instruments, eval = FALSE}
result <- run_mr(
  exposure = exposure_data,
  exposure_id = "PCSK9",
  outcome = outcome_data,
  outcome_id = "CHD",
  instruments = c("rs11591147", "rs2479409", "rs11583680")
)
```

By default, instruments missing from the exposure data produce a warning but
the analysis continues with the available SNPs. Set `instruments_strict = TRUE`
to error instead:

```{r manual-strict, eval = FALSE}
result <- run_mr(
  exposure = exposure_data,
  exposure_id = "PCSK9",
  outcome = outcome_data,
  outcome_id = "CHD",
  instruments = c("rs11591147", "rs2479409"),
  instruments_strict = TRUE
)
```

## Interpreting MR Results

<!-- TODO: print/summary methods -->
<!-- TODO: F-statistics and what they mean -->
<!-- TODO: Steiger filtering -->
<!-- TODO: methods_skipped -->

## Colocalization

### Quick colocalization (ABF only)

<!-- TODO: run_coloc(..., methods = "abf") -->

### Full colocalization (SuSiE + signals)

<!-- TODO: run_coloc() full example -->

### Case-control outcomes

<!-- TODO: outcome_type = "cc", outcome_s = ... -->

## Two-Stage Batch Workflow

<!-- TODO: Stage 1 (IVW screen, all proteins) → Stage 2 (sensitivity + coloc, p < 0.05) -->
<!-- TODO: reference run_decode_iop_pipeline.R pattern -->
