---
title: "mrpipeline Developer Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mrpipeline Developer Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Package Architecture

<!-- TODO: diagram/description of run_mr() → helpers → TwoSampleMR/ieugwasr -->
<!-- TODO: diagram/description of run_coloc() → helpers → coloc/SuSiE -->

## S3 Classes

### `mr_result`

<!-- TODO: full structure documentation (instruments, f_stats, steiger, methods_skipped, ld_matrix, params) -->

Every `run_mr()` call returns an `mr_result` object — even when no instruments
survive filtering. The `status` field indicates success or failure:

- `"success"` — analysis completed normally
- `"no_instruments"` — no instruments survived filtering/clumping/exclusion
- `"no_harmonised_variants"` — harmonisation removed all variants
- `"singular_ld_matrix"` — `ld_correct = TRUE`, the GLS weight matrix is
  singular, and an LD-correctable method was requested

The `status_reason` field provides a human-readable explanation (e.g.
`"No significant instruments in cis region for 'PCSK9'"`).

#### `$instruments` vs `$harmonisation`

Two frames, deliberately not one (GitHub issue #17):

- `$instruments` -- the kept variants, filtered to `mr_keep == TRUE` and
  deduplicated. What the MR estimates are computed from.
- `$harmonisation` -- the complete, unfiltered `harmonise_data()` output:
  every candidate variant and every column, including the `mr_keep`,
  `palindromic`, `ambiguous` and `remove` flags that say *why* a variant did
  or did not survive.

Before this, the flags were discarded inside `harmonise_and_filter()` and
there was no way to ask what harmonisation had done short of running it
again. `harmonise_and_filter()` now returns `list(data =, raw =)` rather than
a bare frame; both callers unpack it immediately.

`$harmonisation` is populated on the `"no_harmonised_variants"` early return
too -- a run that harmonised down to nothing is precisely when someone needs
to see which flag cost each variant its place.

#### `$results` data frame

The `$results` data frame has one row per MR method and the following columns:

| Column | Description |
|--------|-------------|
| `exposure` | Exposure label (from `exposure_id`) |
| `outcome` | Outcome label (from `outcome_id`) |
| `method` | Method label, from the registry's `label` column (`mr_methods()$label`); raw TwoSampleMR passthrough rows carry TwoSampleMR's own label |
| `nsnp` | Number of instruments used |
| `b` | Effect estimate (log-OR scale for case-control outcomes) |
| `se` | Standard error of `b` |
| `pval` | P-value |
| `ld_corrected` | Logical: whether the LD matrix was used for this fit. `TRUE` only for `ivw_random`, `ivw_fixed` and `egger` under `ld_correct = TRUE`; every other row is `FALSE` |
| `model` | `"random"`, `"fixed"`, or `NA` where the fixed/random distinction does not apply (weighted median, PRESSO, ConMix, Wald ratio, passthrough) |
| `lo_ci` | Lower 95% CI on the log-OR scale (`b - 1.96 * se`) |
| `up_ci` | Upper 95% CI on the log-OR scale (`b + 1.96 * se`) |
| `or` | Odds ratio (`exp(b)`) |
| `or_lci95` | Lower 95% CI for OR (`exp(lo_ci)`) |
| `or_uci95` | Upper 95% CI for OR (`exp(up_ci)`) |

Every row is built by `mr_result_row()` from its registry entry, so the
schema cannot differ between dispatch branches (which would make the final
`do.call(rbind, results_list)` fail). The OR columns are added by
`TwoSampleMR::generate_odds_ratios()` in the results assembly step on a
successful run; early returns carry the nine base columns of
`empty_mr_results()`, the `results` default of `new_mr_result()`. For
quantitative outcomes the OR columns are still computed but should be
interpreted cautiously -- they reflect exponentiated effect sizes, not true
odds ratios.

#### The method registry and LD correction

`R/mr_methods.R` holds `mr_method_registry()`, the single source of truth for
what `run_mr()` can run. `run_mr()` reads everything method-related from it:
the valid shortcut names, which raw TwoSampleMR names are refused because a
shortcut already wraps them (derived from the `engine` column), the minimum
instrument count behind every `"Requires >= n instruments"` skip reason, the
`label`/`model` written into each `$results` row, and which methods warn
`has no LD-corrected form` when `ld_correct = TRUE`. `mr_methods()` exports
it, `rd_method_table()` renders it into `?run_mr` through a roxygen `@eval`
tag, and the user guide renders it with knitr. Before the registry the
shortcut list was hard-coded in three places that had drifted apart, which
is how `ivw_fe` came to be half-wired (GitHub issue #27).

```{r method-registry-full}
tab <- mrpipeline::mr_methods(detail = "full")
# Backticks keep MathJax from reading `$results` ... `$steiger` as maths.
tab$output <- paste0("`", tab$output, "`")
knitr::kable(tab)
```

`ld_correctable` is declared, not inferred -- nothing inspects upstream
function signatures at run time. It is kept honest by `engine_ld`: a row may
claim LD support only by naming the function that provides it, and
`test-mr_methods.R` asserts `ld_correctable == !is.na(engine_ld)`. There is
deliberately no free-text notes column, because notes about upstream
internals go stale silently when a dependency changes; `engine`/`engine_ld`
name what mrpipeline calls and so stay true as long as the dispatch does.

The LD path itself: `compute_ld_matrix()` and `align_to_ld_matrix()` run
once, before dispatch, and the aligned matrix goes into a single
`MendelianRandomization::mr_input(correlation = ld_mat)` that every corrected
fit shares. The matrix -- not the `correl = TRUE` flag -- is what carries the
correction; `mr_ivw()`/`mr_egger()` read it from that object. The IVW model
is pinned explicitly (`model = "random"` / `"fixed"`) because
`MendelianRandomization`'s `"default"` is fixed effects below 4 instruments
and random above, which would make `ivw_random` mean two different
estimators depending on how many SNPs survived clumping. Egger needs no
pinning: `mr_egger()` has no `model` argument and is multiplicative random
effects on both paths (`/min(1, sigma)` in TwoSampleMR, `*max(1, rse.corr)`
in MendelianRandomization). If alignment drops every instrument, `run_mr()`
returns `status = "no_harmonised_variants"` rather than reaching the GLS fit
with zero rows.

Once `ld_input` exists, `run_mr()` also checks `rcond()` of the GLS weight
matrix `O = diag(se_y) R diag(se_y)` (`gls_weight_matrix()`) and warns below
`1e-10`: every LD-corrected fit solves that matrix, so a near-singular one
-- identical or near-identical instruments, or more instruments than panel
individuals -- makes all of them unstable, and before this it happened
silently.

Below that threshold the fits do not merely become unstable: on an exactly
singular matrix `MendelianRandomization`'s own `solve()` aborts, and the
error escaped `run_mr()` as a bare `simpleError` naming `solve.default(omega)`
(issue #36). So when the matrix is singular *and* at least one
LD-correctable method was requested -- `registry$ld_correctable`, which
covers the correlated diagnostics as well, since they solve the same matrix
-- `run_mr()` returns `status = "singular_ld_matrix"` before any fit, with
`collinear_pairs()` naming the `|r| >= 0.999` pairs in `status_reason`. With
no LD-correctable method requested nothing solves the matrix, so the run
continues on the warning alone.

**The diagnostics are corrected too** (issue #31). `#28` exempted them on
the grounds that they produce no `$results` row, which left `$heterogeneity`,
`$pleiotropy` and `$loo` computed by TwoSampleMR on the uncorrected data with
nothing on the object saying so -- a caller reading Cochran's Q off an
LD-corrected result got the naive Q. Each frame now carries an
`ld_corrected` column on both arms, and under `ld_correct = TRUE` the values
come from the correlated fits, keeping TwoSampleMR's column names and row
order so downstream code and `plot(x, type = "loo")` are unaffected:

| frame | uncorrected arm | LD-corrected arm |
|---|---|---|
| `$heterogeneity` | `TwoSampleMR::mr_heterogeneity()` | `heterogeneity_correlated()`: `mr_ivw(correl = TRUE)@Heter.Stat` (`n - 1` df) and `mr_egger(correl = TRUE)@Heter.Stat` (`n - 2` df, `n >= 3`) |
| `$pleiotropy` | `TwoSampleMR::mr_pleiotropy_test()` | `pleiotropy_correlated()`: `mr_egger(correl = TRUE)@Intercept` / `@StdError.Int` / `@Pvalue.Int` |
| `$loo` | `TwoSampleMR::mr_leaveoneout()` | `loo_correlated()`: per-SNP random-effects GLS refit (below) |

`$heterogeneity$method` keeps TwoSampleMR's labels (`"Inverse variance
weighted"`, `"MR Egger"`) on both arms: Q is model-agnostic, so neither
`$results` IVW label would be right, and a stable label makes the column the
only difference. Steiger filtering is the one diagnostic left alone -- it
compares per-SNP r^2 values and has no weight matrix -- so it neither warns
nor carries the column.

`loo_correlated()` is the one piece with no upstream implementation:
`MendelianRandomization::mr_loo()` takes no `correl` argument and never reads
the correlation slot. It is deliberately **not** n refits of `mr_ivw()`,
which is n solves of an (n-1)x(n-1) system -- O(n^4), measured at 15 s for
n = 300 and minutes for n = 600. Instead it inverts `O` once and, for each
SNP `i`, obtains the inverse with that row and column removed by the
Schur-complement identity `W[-i, -i] - W[-i, i] W[i, -i] / W[i, i]`
(O(n^2) per SNP, O(n^3) total; 0.12 s at n = 300, 1 s at n = 600; agrees
with the refit loop to machine precision), then reproduces `mr_ivw()`'s
random-effects arithmetic exactly -- residual SE on `m - 1` df, floored at 1,
normal p-value. Two guards: below `rcond_min` it falls back to direct per-SNP
solves, and `test-ld-diagnostics.R` pins every row to a literal
`mr_ivw(correl = TRUE)` refit, so a change in upstream's convention fails CI
rather than drifting silently.

#### Egger intercept / pleiotropy test

`TwoSampleMR::mr_pleiotropy_test()` (or `pleiotropy_correlated()` under
`ld_correct = TRUE`) is run automatically whenever `"egger"` is in `methods`
and there are >= 3 instruments. The result is stored in `$pleiotropy` as a
data frame with columns `egger_intercept`, `se`, `pval`, `exposure`,
`outcome`, `ld_corrected`. The `"pleiotropy"` shortcut in `methods` triggers
the same test independently (i.e. without also running the Egger slope
estimate).

#### Heterogeneity test (Cochran's Q)

`TwoSampleMR::mr_heterogeneity()` (or `heterogeneity_correlated()` under
`ld_correct = TRUE`) runs when `"heterogeneity"` is in `methods` and there
are >= 2 instruments; otherwise `methods_skipped["heterogeneity"]` records
the reason. The result is stored in `$heterogeneity` as a data frame with one
row per method (MR Egger when there are >= 3 instruments, then IVW),
columns `Q`, `Q_df`, `Q_pval`, `exposure`, `outcome`, `ld_corrected`.

#### Leave-one-out analysis

`TwoSampleMR::mr_leaveoneout()` (or `loo_correlated()` under
`ld_correct = TRUE`) runs when `"loo"` is in `methods` and there are >= 3
instruments (below that, dropping one SNP just reproduces the remaining
SNP's Wald ratio, which isn't informative); otherwise
`methods_skipped["loo"]` records the reason. The result is stored in `$loo`
as a data frame with one row per SNP plus a pooled `"All"` row, columns
`SNP`, `b`, `se`, `p`, `exposure`, `outcome`, `ld_corrected`.

#### Flipping exposure direction -- deliberately not provided

`run_mr()` has no `flip_beta`/`exposure_label` argument, and none should be
added. An earlier design (2026-07-29 to 2026-08-01) negated
`harmonised$beta.exposure` once, immediately after harmonisation, so the
flip propagated consistently into `$results`, `$instruments`, and `$loo`.
It was removed because a shared-library flag that only ever touches
`beta.exposure` -- never `effect_allele.exposure`/`other_allele.exposure`/
`eaf.exposure`/`beta.outcome` -- hides that scope from the caller, inviting
misuse. Projects that need to redefine exposure direction (e.g. modelling a
drug's inhibition mechanism) should write their own `flipped_beta()`-style
helper in project-level code and apply it to the exposure/harmonised data
before it reaches `run_mr()`. See the `mrpipeline`/root `CLAUDE.md` files
for the full rationale.

#### `print.mr_result()` and `summary.mr_result()`

`print.mr_result()` shows a one-line summary using the first result row,
including the OR and 95% CI (when available), tagged `[LD-corrected]` when
that row's `ld_corrected` is `TRUE`. `summary.mr_result()` shows all methods
with OR columns -- each tagged with its effects model where the label does
not already carry it (so `MR Egger [random effects]`) and, on an LD-corrected
run, `[LD-corrected]` or `[not LD-corrected]` -- then F-statistics, Steiger
result, the Egger intercept (when computed), the heterogeneity (Cochran's Q)
table (when computed), a leave-one-out row count (when computed), skipped
methods, and finally an **LD correction** section listing the methods the
matrix was and was not applied to. That section is keyed off
`$results$ld_corrected`, not off the presence of `$ld_matrix`, so a run whose
methods all lacked a correlated form reads "Applied to: none" rather than
being announced as LD-corrected.

### `coloc_result`

Every `run_coloc()` call returns a `coloc_result` object — even when analysis
cannot proceed (e.g. no SNPs in region). The object contains:

- `coloc_abf` — output of `coloc::coloc.abf()`, or `NULL`
- `coloc_susie` — output of `coloc::coloc.susie()`, or `NULL`
- `coloc_signals` — output of `coloc::coloc.signals()`, or `NULL`
- `coloc_prop_test` — output of `colocPropTest::coloc.prop.test()`, or `NULL`
- `n_snps` — integer, number of SNPs used in the analysis
- `harmonised_data` — the harmonised data the analysis actually ran on:
  filtered, then subset and reordered by `align_to_ld_matrix()` so its rows
  correspond one-to-one with the coloc datasets
- `harmonisation` — the complete, unfiltered `harmonise_data()` output,
  with the `mr_keep`/`palindromic`/`ambiguous`/`remove` flags
- `methods_skipped` — named character vector (method name → reason skipped).
  `susie` is recorded both when SuSiE fails and when it runs but finds no
  credible set for one of the traits (`"no credible sets in outcome"`), since
  `coloc.susie()` then has nothing to pair; in the latter case the SuSiE fits
  are kept and `coloc.signals()` still runs on them
- `params` — list of all input parameters
- `status` — one of `"success"`, `"no_snps_in_region"`, `"too_few_snps"`,
  `"no_harmonised_variants"`
- `status_reason` — human-readable explanation when `status != "success"`

`print.coloc_result()` shows a one-line summary: N SNPs, ABF PP.H4, and
SuSiE max PP.H4 with credible set count (if available).

`summary.coloc_result()` shows full details: all posterior probabilities
(H0–H4), SuSiE credible set pairs, signals hits, and skipped methods.

### Plot methods

Both S3 classes have `plot()` methods defined in `R/plot.R`:

- `plot.mr_result(x, type)` — wraps TwoSampleMR plotting functions, each of
  which returns a **list** of ggplots keyed by exposure/outcome ID (one per
  pair; the outcome ID is random, so index with `[[1]]` rather than by name):
  - `"scatter"` (default): `TwoSampleMR::mr_scatter_plot()`
  - `"forest"`: `TwoSampleMR::mr_forest_plot()`
  - `"funnel"`: `TwoSampleMR::mr_funnel_plot()`
  - `"loo"`: `TwoSampleMR::mr_leaveoneout_plot()` on `x$loo`
- `plot.coloc_result(x, type)` — custom ggplot2 plots, except `"locuszoom"`:
  - `"pp_bar"` (default): bar chart of ABF posterior probabilities (H0–H4)
  - `"regional"`: side-by-side regional association plots (-log10(p) vs
    position) for exposure and outcome
  - `"locuszoom"`: LD-coloured regional plots via the (`Suggests`-only)
    `locuszoomr` package, dispatched to the internal
    `plot_coloc_locuszoom(x, ens_db, bfile, plink_bin, index_snp, ...)`.
    Unlike every other plot function in this file, this one does **not**
    return a ggplot object (or any re-renderable object) -- `locuszoomr`'s
    `locus()`/`multi_layout()` draw directly to the current graphics device
    as a side effect, like base/grid plotting functions, and
    `plot_coloc_locuszoom()` returns `NULL` invisibly regardless of success.
    Callers must open a device (`pdf()`, `png()`, ...) before calling and
    `dev.off()` after, to save the result. `ens_db`, if given as a string,
    must already be **attached** via `library()` -- `locuszoomr::locus()`
    resolves it via `get(ens_db)` on the search path and errors if it's
    only installed, not attached. LD (r2) is computed against a single
    index SNP (default: the SNP with the lowest `pval.exposure`, matching
    `locuszoomr::locus()`'s own default) via the internal helper
    `compute_ld_to_index()` (see below), which reuses `compute_ld_matrix()`
    -- no new LD mechanism or network calls. There is no genome-build field
    stored on `coloc_result`, so the caller must supply `ens_db` matching
    whatever build the data/`bfile` actually use (e.g. `EnsDb.Hsapiens.v75`
    for GRCh37, `EnsDb.Hsapiens.v86` for GRCh38) -- no auto-detection.

Both S3 methods return `NULL` invisibly for non-success results with an
informative message.

### forest_plot(), outcome_forest_plot(), and table_forest_plot()

`R/plot.R` also defines three plain exported functions (not S3 methods --
none of `forest_plot()`'s list-of-`mr_result` input, `outcome_forest_plot()`'s
and `table_forest_plot()`'s caller-built data frame input fit `plot(x, ...)`
dispatch):

- `forest_plot(results, methods, relabel, ...)` -- one row per *method*.
  Filters and reorders each result's `$results` to `methods`, binds them
  (list order becomes section order when `results` is a named list), then
  calls `TwoSampleMR::forest_plot_1_to_many()` with `TraitM = "method"`.
  Default `methods` puts `"IVW (fixed effects)"` above `"IVW (random
  effects)"`. Labels no longer encode LD correction (that is the
  `ld_corrected` column), so LD-corrected results match the same defaults --
  previously every `"(LD-corrected)"` row was silently dropped. The
  `relabel` argument (empty by default) is applied to the combined data's
  `method` column *after* filtering/ordering, so it never affects matching
  against the raw labels `run_mr()` produced.
- `outcome_forest_plot(mr_res, xlab, method, colour_by, shape_by, ...)` --
  one row per *outcome*. Standalone `ggplot2` code rather than a
  `forest_plot_1_to_many()` wrapper, because that function force-blanks its
  `exposure` column internally and so cannot map colour or shape to a
  grouping variable. `method` is a plain inclusion filter (not a
  priority/fallback selector): pass one label for one model, or both to get
  two rows for any outcome where both were computed. `colour_by`/`shape_by`
  each map an arbitrary column in `mr_res` to that aesthetic, so e.g.
  instrument type and model (fixed/random effects) can be shown at once
  without colliding.

Implementation note for anyone extending `outcome_forest_plot()`: the
mapping always includes `group = .data$method`, even when `method` isn't
mapped to colour or shape. This is what makes `ggplot2::position_dodge()`
separate two rows that share the same `outcome` (e.g. a fixed-effects and a
random-effects estimate for the same outcome) instead of fully overplotting
them -- removing that `group` binding would silently break the dodge for
callers who only map `colour_by`/`shape_by` to something other than
`method`.

- `table_forest_plot(dat, null_value, xlab, box_colour)` -- one row per
  outcome, rendered as an inline result/CI/p-value table alongside the
  plotted estimate, via the (`Suggests`-only) `forestplot` package rather
  than `ggplot2`. Ported from a project-level script
  (`NLRP3/Lanyue_forestplot_script/build_cardiometabolic_forestplot.R`) that
  was a deliberate one-off exception to `outcome_forest_plot()`'s usual
  convention. Unlike that original script, this function never opens its
  own graphics device (no internal `pdf()`/`dev.off()`) -- it returns the
  `forestplot` object, matching how every other plot function in this file
  behaves; the caller renders it via `plot()`/`print()` and controls the
  output device.

## Exported Utility Functions

### Single-cell RNA eQTL format functions

Four functions format raw single-cell eQTL datasets into the TwoSampleMR
exposure format expected by `run_mr()` and `run_coloc()`. They live in
`R/format_single_cell_onek1k.R` and `R/format_sc_RNA_eqtl.R`. All functions:

- Produce a TwoSampleMR exposure data frame with `chr.exposure`, `pos.exposure`,
  and `eaf.exposure` (where available) so that `run_coloc()` can perform its
  window filter and MAF computation without requiring extra arguments.
- Use the phenotype label convention `GENE___cell_type` (triple underscore),
  matching what `run_mr()` and downstream splitting code expect.
- Accept a `cell_type` argument that, when `NULL`, is inferred from the filename.

#### `format_single_cell_onek1k()`

Handles the OneK1K combined eQTL table (`esnp_table.tsv.gz` or per-cell-type
files). Key column mapping:

| Raw column | Role | Notes |
|---|---|---|
| `A2` | effect allele | Assessed allele; `A2_FREQ_ONEK1K` is its frequency |
| `A1` | other allele | |
| `SPEARMANS_RHO` | beta | Used as a standardised effect size |
| derived | se | `|rho| / qnorm(P_VALUE / 2, lower.tail = FALSE)` |
| `CHR`, `POS` | chr, pos | Required by `run_coloc()` window filter |
| `A2_FREQ_ONEK1K` | eaf | Required by `run_coloc()` MAF computation |

When the file contains a `CELL_ID` column (multi-cell-type combined file),
rows are filtered to the requested cell type. An optional `sample_size`
argument stores `samplesize.exposure` for `run_coloc()`.

#### `format_sceqtl_1m_scbloodnl()`

Handles per-cell-type files from the 1M-scBloodNL cohort
(`*_expression_eQTLsFDR-ProbeLevel.txt.gz`). Beta and SE are both derived
because the raw files report Z-scores rather than effect sizes:

- `N_total = sum(DatasetsNrSamples split on ";")`
- `se = 1 / sqrt(N_total)`
- `beta = OverallZScore * se`
- `other_allele` derived by splitting `SNPType` on "/" and taking the allele
  that differs from `AlleleAssessed`

Rows where the other allele cannot be determined (allele not found in
`SNPType`) are dropped with an informational message. Filtering to cis eQTLs
(`CisTrans == "cis"`) is applied by default via the `cis_only` argument. Note
that EAF is not available in these files, so `eaf.exposure` will be absent
from the output, which means `run_coloc()` will need an explicit `exposure_n`
argument and sample size cannot be inferred from the formatted data.

#### `format_sceqtl_dice()`

Handles per-cell-type VCF files from the DICE database. VCF meta-information
lines (`##`) are skipped by the internal `read_vcf_data()` helper, which counts
comment lines via a streaming connection before calling `data.table::fread()`.
The INFO field is parsed with regex to extract `Gene`, `GeneSymbol`, `Pvalue`,
and `Beta`. SE derivation: `|Beta| / sqrt(qchisq(Pvalue, df = 1))`.

Key filtering decisions:
- Only biallelic SNPs (single nucleotide, A/C/G/T bases) are retained.
- Palindromic SNPs (A/T and C/G) are **dropped** because DICE provides no
  allele frequency data to resolve strand ambiguity.
- `ALT` is treated as the effect allele; `REF` as the other allele.

No EAF or sample size is available in the DICE VCF format.

#### `format_sceqtl_dynamic_cseqtl()`

Handles per-cell-type pre-formatted files (`*_500kb_combined.MR.tsv.gz`).
These files already contain standard columns (`SNP`, `beta`, `se`,
`effect_allele`, `other_allele`, `pval`, `chr`, `pos`), so the function's
only job is to append the cell type to the gene name and call
`TwoSampleMR::format_data()`. No EAF or sample size is available.

#### Internal helper: `read_vcf_data()`

Lives in `R/format_sc_RNA_eqtl.R`, tagged `@keywords internal`. Opens a
streaming connection to the file, counts `##` comment lines one at a time
(memory-efficient), reads the `#CHROM` header line to extract column names,
then calls `data.table::fread()` with `skip` and `col.names`. Works for both
plain and gzip-compressed VCF files.

### `get_gene_coords()`

Defined in `R/get_gene_coords.R`. Queries Ensembl via biomaRt for gene
coordinates. Key design decisions:

- Uses `biomaRt::useEnsembl()` with `host = "grch37.ensembl.org"` for GRCh37
  or the default host for GRCh38
- Filters to standard chromosomes (1–22, X, Y)
- Deduplicates by gene + chromosome (widest range), preferring autosomes
- Warns about genes not found in Ensembl

### `last_allele_check()`

Defined in `R/last_allele_check.R`. Returns the record stored by the most
recent `check_allele_orientation()` call (the `httr2::last_response()`
pattern), or `NULL` with a message if none has run. The record is a named list
-- `status` (`"pass"`/`"fail"`/`"skipped"`), `n`, `n_complementary`, `prop`,
`threshold`, `min_n`, `allele_check`, `exposure`, `outcome`, `id.exposure`,
`id.outcome`, `n_sampled`, and `variants`, a data frame with one row per SNP that had both
frequencies (palindromic and `remove` rows included but flagged, with a
`score` column to sort offenders). A list rather than a data frame with
attributes so the summary fields survive `dplyr` verbs and `[`, and no S3
class or print method is needed.

The storage is a package-level environment, `.mrpipeline_env <-
new.env(parent = emptyenv())` in `R/zzz.R` (the same idiom as codeminer's
`.codeminer_env`). The namespace is locked on load but bindings inside that
environment are not, so helpers can assign `.mrpipeline_env$last_allele_check`
at run time. It is the package's only mutable state.

## Internal Helper Functions

All helpers live in `R/helpers.R`, are tagged `@keywords internal`, and are not exported.

### `harmonise_and_filter()`

Single choke point for harmonisation, used by both `run_mr()` and
`run_coloc()`. Wraps `TwoSampleMR::harmonise_data()`, runs
`check_allele_orientation()` on the raw output, returns a 0-row frame if
harmonisation produced nothing usable (so callers hit their `nrow == 0`
early-return instead of erroring on a missing `mr_keep` column), then filters
to `mr_keep == TRUE` and drops duplicate SNPs. The allele check runs *before*
the `mr_keep` filter so SNPs later dropped for a missing `beta`/`se` still
contribute their frequencies, and on every call so the stored record is never
stale. `run_mr()` passes `check = FALSE` because it has already checked a
larger set via `check_allele_orientation_gwas()`; `run_coloc()` keeps the
default because its window is already large.

`harmonise_and_filter()` returns `list(data =, raw =)`: `data` is the filtered
frame the analysis runs on, `raw` the complete unfiltered `harmonise_data()`
output that the result objects store as `$harmonisation` (issue #17). A bare
frame plus an attribute was the alternative, and was rejected because dplyr
operations drop attributes silently -- the list makes the second frame
impossible to lose by accident.

`harmonisation_summary()` reduces `raw` to the counts both `summary()` methods
print. Its one subtlety is that the reason counts *overlap* -- every ambiguous
variant is by definition palindromic -- so they are listed, never summed into
a breakdown of `n_dropped`. Which flags actually cost a variant its place also
depends on the action level (`remove` always, `ambiguous` from 2, `palindromic`
only at 3), and `n_incomplete` covers variants `harmonise_data()` drops for
missing beta/se, which none of the three allele flags reveals.

`action` (exposed as `harmonise_action` on both entry points, validated by
`validate_harmonise_action()` and recorded in `params`) passes straight
through to `harmonise_data()`. Only palindrome handling differs between the
three levels -- `1` skips the frequency-based flip entirely, `2` applies it,
`3` sets `mr_keep = FALSE` for every palindromic, ambiguous or incompatible
SNP so `harmonise_and_filter()`'s existing `mr_keep` filter drops them. The
letter-based alignment of non-palindromic variants happens at all three
levels, which is why `check_allele_orientation()`'s verdict does not depend
on `action` (there is a test asserting the swapped-allele fixture still fails
at 1, 2 and 3). `check_allele_orientation_gwas()` takes the same `action` so
the check describes the harmonisation the analysis will actually use.

`validate_harmonise_action()` deliberately does *not* coerce to integer.
`harmonise_data()` embeds `action` as a column in its output, so `2` -> `2L`
would leave mrpipeline's harmonised frame differing from a plain
`harmonise_data()` call by that column's storage mode alone -- which a test
asserts against. It does reject vectors: TwoSampleMR accepts one action per
outcome, but both entry points handle exactly one outcome, so a vector is a
mistake rather than a recycling opportunity.

### `check_allele_orientation()`

The verdict behind `allele_check` (GitHub issue #18). Some GWAS files use
`A1`/`A2` to mean REF/ALT with `BETA`/`FRQ` per `A2`; `format_gwas()` reads
`A1` as the effect allele, so every beta is silently inverted. Harmonisation
is per-variant and letter-based, so it faithfully applies the swapped labels
and the output is internally consistent -- the only trace is that
`eaf.outcome` ends up describing the other allele, i.e. `eaf.outcome ~ 1 -
eaf.exposure` across the set. The helper counts, among *informative* rows
(both EAFs present, `palindromic == FALSE`, `remove == FALSE`, one row per
SNP), how many have `eaf.exposure` closer to `1 - eaf.outcome` than to
`eaf.outcome`; it fails when that proportion exceeds 0.70 with at least 10
informative rows, and is `"skipped"` below that. Ties (`eaf.outcome == 0.5`)
count as not complementary; near-0.5 frequencies can only dilute the
proportion, never trigger it.

Palindromic SNPs are excluded because TwoSampleMR resolves their strand *from*
the frequencies: when effect allele and frequency are both mis-assigned the
two flips cancel, `eaf.outcome` looks consistent, and the harmonised beta can
be identical to the correct value while every non-palindromic beta in the same
set is inverted (the tests prove this on the fixture). Including them would
only dilute the statistic -- and once the check fails, every palindromic
strand call in that pair is unreliable regardless of its beta.

The check is symmetric: it cannot say which dataset is wrong. The condition
message (classes `mrpipeline_allele_check_error` /
`mrpipeline_allele_check_warning`) lists the five worst offenders inline, the
tie-breakers (reference panel frequencies, a known-direction variant,
provenance) and the `col_map` fix. The record is stored in `.mrpipeline_env$last_allele_check`
in every mode, including `"none"`.

A `"skipped"` verdict -- no usable frequencies, or fewer than `min_n`
informative rows -- warns (class `mrpipeline_allele_check_unverified`) rather
than informing. It used to be a `cli_inform()` gated behind `verbose`, which
put "could not be checked" at the same level as ordinary progress output:
scanning a log, silence reads as a pass, and in a real run two unchecked pairs
slipped past that way while every other harmonisation passed (GitHub issue
#21). The warning is emitted regardless of `verbose`, since `verbose` gates
informational output only; `allele_check = "none"` suppresses it, on the
grounds that a user who has switched the check off does not need telling it
did not run. The message points at `?format_gwas`, section *If the check
cannot run*, which documents the frequency-free fallback: anchoring on a
variant with an established effect direction.

### `check_allele_orientation_gwas()`

`run_mr()`'s instrument set is often below the 10-SNP minimum (a cis-MR may
have three instruments), but `run_mr()` holds the full exposure and outcome
before it narrows the outcome to instrument rsIDs. This helper harmonises the
instruments plus up to 1000 further shared SNPs and hands the result to
`check_allele_orientation()`. The extra SNPs are evenly spaced picks through
the sorted shared rsIDs: deterministic (reproducible, and the caller's RNG
state is untouched) and effectively random with respect to genomic position.
Cost scales with the total rows of the two inputs, not with the number of
instruments, and is recorded as `timing[["allele_check"]]`. Profiled on a
42M-row genome-wide exposure against a 12M-row genome-wide outcome (9.6M
shared rsIDs): `intersect()` ~3 s, `setdiff()` ~0.6 s, the radix sort of
the shared rsIDs ~1.8 s, the two `%in%` subsets ~1.2 s, formatting and
harmonising the ~1000 sampled SNPs a few milliseconds -- about 6 s in all.
A 200k-SNP exposure takes well under a second.

The sort is `sort(shared, method = "radix")` deliberately (issue #40). Base
`sort()` chooses radix only for numeric, factor and logical vectors; for
character vectors it falls back to shell sort with a locale collation call
per comparison, which took ~50 s on the same 9.6M rsIDs -- 90% of the check
and ~95% of a cis-MR's runtime in a pipeline that passed genome-wide
inputs. Radix orders by byte (C locale) rather than by `LC_COLLATE`, so the
sample is now also independent of the session locale; for rsIDs the two
orders are identical, and a test pins the locale independence.

Caching the check across `run_mr()` calls that share an exposure/outcome
pair was considered and rejected: hashing the two rsID columns to build a
key costs ~3 s against the ~5 s it would save, hashing the full frames
costs more than the check itself, keying on `exposure_id`/`outcome_id` is
unsafe (the same object is routinely passed under different ids), and a
caller-supplied precomputed verdict is a skip switch by another name. The
check therefore runs on every call, which also keeps the
`last_allele_check()` record honest.


### `compute_ld_matrix()`

Runs PLINK `--r square` (via `ieugwasr::ld_matrix()`) on the local `bfile`
for a vector of rsIDs and returns `list(ld =, alleles =)`: the **signed**
correlation matrix with rsID-only dimnames, and the panel's `ld_a1`/`ld_a2`
allele coding for each SNP. The sign matters -- a pair in repulsion enters
the GLS as a negative `r` -- and it is anchored to the panel's A1 allele, so
the allele table is returned alongside rather than discarded. `bfile` is
always required: there is no API fallback, because the matrix must be
computed on exactly the instrument set and allele coding that
`align_to_ld_matrix()` then reconciles with the exposure.

### `compute_ld_to_index()`

Reuses `compute_ld_matrix()` and squares its signed correlation column for
one chosen index SNP to get r2 -- no separate LD mechanism. Backs
`plot_coloc_locuszoom()`'s LD colouring; not used anywhere else.

### `clump_instruments()`

<!-- TODO: document pseudo-p-value approach, API vs local mode, session-level message suppression -->

### `align_to_ld_matrix()`

Intersects the harmonised instruments with the matrix, reorders both to the
shared SNPs, and re-orients the matrix's sign to the exposure's effect
allele: a SNP whose `effect_allele.exposure`/`other_allele.exposure` are the
panel's `ld_a2`/`ld_a1` gets its row and column sign flipped; a SNP whose
alleles match neither orientation, or that is palindromic with an ambiguous
frequency (0.42-0.58 in either dataset), is dropped. This is why an
LD-corrected `run_mr()` can end up with fewer instruments than the same call
uncorrected, and why one `mr_result` is always one instrument set under one
weight matrix. Returns `list(data =, ld_matrix =, ld_sign =)`; `run_mr()`
takes `data` as its new `harmonised` frame and builds the shared
`MendelianRandomization::mr_input()` from `ld_matrix`.

### `plink_option()`

Resolves PLINK resource limits (threads, memory) from R options or environment
variables. Checks `getOption("mrpipeline.plink_{param}")` first, then falls back
to the `MRPIPELINE_PLINK_{PARAM}` environment variable. Returns `NULL` if
neither is set (PLINK auto-detects). Used as the default value for
`plink_threads` and `plink_memory` in `run_mr()` and `run_coloc()`.

### `fread_file()`, `check_gz_support()`

Every user-supplied file path in the package is read via `fread_file()`, a thin
wrapper around `data.table::fread()` that first calls `check_gz_support()`.

GWAS summary statistics are distributed gzipped as a matter of course, but
`data.table` lists `R.utils` -- the package `fread()` needs to decompress
`.gz`/`.bz2` -- only in `Suggests`. That made a documented capability depend on
an undeclared package: a downstream project hit
`"To read gz files directly, fread() requires 'R.utils'"` on a clean machine
because nothing in the dependency chain pulled it in (issue #20). `R.utils` is
therefore in `mrpipeline`'s **`Imports`**, which also means `renv` captures it
in every downstream lockfile automatically. Do not move it back to `Suggests`.

`check_gz_support()` is the backstop for a broken library rather than an
expected failure mode: it aborts with an actionable `cli` message at the top of
the call stack instead of deep inside `fread()`, which matters for a pipeline
that may already have spent minutes on earlier targets. Its `has_rutils`
argument (`NULL` by default, meaning "look it up with `requireNamespace()`, but
only for a compressed path") exists so tests can exercise the failure branch.

### `eaf_to_maf()`, `resolve_sample_size()`

<!-- TODO: brief documentation for each -->

## Code Conventions

### Messages, Warnings, and Errors

Always use the `cli` package. Never use `message()`, `warning()`, or `stop()` directly.

```r
cli::cli_inform("Processing {protein}...")      # informational (goes to stderr)
cli::cli_warn("Only {n} SNP(s) available.")     # warning
cli::cli_abort("bfile is required for coloc.")  # error (stops execution)
```

The `verbose` parameter is a **planned future feature**. When implemented, all `cli::cli_inform()` calls should be gated behind it.

### Pipe

Use the native `|>` pipe exclusively. Never use `%>%` from magrittr.

### String Operations

Prefer `stringr` functions over base R equivalents:

```r
# Good
stringr::str_remove(x, "_.*")
stringr::str_detect(x, "^chr")
stringr::str_replace(x, "chr", "")

# Avoid
gsub("_.*", "", x)
grepl("^chr", x)
sub("chr", "", x)
```

### Namespace

Prefer `pkg::fn()` over `@importFrom`. Only add `@importFrom` entries when a function is called very frequently in a hot path.

### Documentation

- All exported functions: `@param`, `@return`, `@examples`, `@seealso`, `@family`
- Internal helpers: `@keywords internal` only; no `@export`
- When modifying an exported function, update its roxygen docs AND this vignette in the same commit

### Vignettes

Vignette output is evaluated, not typed (issue #33): a hand-written `#>` line
is a claim about what the code does, an evaluated chunk is a test of it, and
`R CMD check` builds the vignettes on every CI platform. The conventions, set
up in the `setup` chunk of `mrpipeline-user-guide.Rmd` and
`mrpipeline.Rmd`:

- **Gate, don't hard-code.** Chunks that need PLINK use `eval = has_ld`,
  where `has_ld` checks the bundled panel *and* wraps
  `genetics.binaRies::get_plink_binary()` in `tryCatch()` -- it downloads
  PLINK on first use and errors, rather than returning `NULL`, when it
  cannot. `has_gg`, `has_susie` and `has_forestplot` gate on `Suggests`. A
  platform without them builds the page with code and no output.
- **`message = FALSE` globally.** `run_mr()`/`run_coloc()` narrate progress
  through messages. `print()`/`summary()` on the result classes also write
  through `cli` messages, so chunks that show them need `message = TRUE`.
  PLINK's own console output goes to the process stdout, which knitr does
  not capture -- it reaches the build log, never the page.
- **`run_coloc()` chunks use `results = "hide"`**, because `coloc` prints its
  posteriors with `print()`; show the result from a separate chunk.
- **Never print a whole `$results`/`$heterogeneity`/`$loo` frame.**
  TwoSampleMR stamps a random `id.outcome` on every run; select columns. For
  the same reason `plot.mr_result()`'s list is indexed with `[[1]]`.
- **Seed anything stochastic** (`set.seed()` visibly in the chunk): the
  weighted median's SE is bootstrapped, and SuSiE is seeded too.
- **Numbers in prose come from the results**, via inline R expressions that
  call `num()` on them (e.g. `num(mr_res$f_stats$min)`); `num()` is defined
  in the setup chunk and prints `[not computed]` when `has_ld` is false. Qualitative claims the prose makes about them ("the
  corrected SEs are smaller", "H1 dominates") are asserted in a hidden
  `include = FALSE` chunk with `stopifnot()`, so the build fails rather than
  publishing a claim the data no longer support.
- **`eval = FALSE` only where it is honest** -- network calls
  (`get_gene_coords()`, API clumping), user-supplied files and large
  downloads -- and say so in the prose; any `#>` output there is labelled
  representative.

## Adding New MR Methods

Add a row to `mr_method_registry()` in `R/mr_methods.R` -- that is the only
way to make a method requestable, and the tests in `test-mr_methods.R` will
fail if the row is inconsistent (a label missing on a `$results`-producing
row, `ld_correctable` claimed without an `engine_ld`, and so on). Then:

1. Add the dispatch branch in `run_mr()`'s method-dispatch block, following
   the existing pattern: `entry <- mr_method_entry("name")`, `too_few(entry)`
   for the skip reason, `warn_no_ld_correction("name")` if `ld_correct` is
   `TRUE` and the method has no correlated form (or fit from the shared
   `ld_input` if it does), and `mr_result_row(entry, ...)` for the row. Never
   build a `data.frame()` by hand -- the row helper is what keeps the
   `$results` schema identical across branches.
2. If the method has a correlated implementation, name it in `engine_ld` and
   set `ld_correctable = TRUE`; if not, leave `engine_ld` as `NA` so the
   warning fires.
3. Nothing else to update: validation, the single-instrument skip list, the
   raw-name shadowing, `?run_mr`'s method table and both vignettes' tables
   all read the registry.

Diagnostics that write to a field other than `$results` (like `steiger` or
`loo`) get `output = "$<field>"`, `label = NA`, and a corresponding argument
on `new_mr_result()`.

## Adding New Coloc Methods

<!-- TODO: where to add in run_coloc() -->
<!-- TODO: how to update coloc_result structure if needed -->

## Test Data

The package ships several bundled datasets (defined in `R/data.R`) and a
minimal LD reference panel. All SNP-level datasets overlap with the LD panel,
so integration tests can run without external downloads.

### Bundled datasets

| Dataset | Rows | Format | Description |
|---------|------|--------|-------------|
| `cd40_sumstats` | 210 | Raw UKB-PPP | CD40 summary statistics (chr6 MHC + chr20 cis), input for `format_pqtl_ukbppp()` |
| `cd40_exposure` | 50 | TwoSampleMR exposure | CD40 chr20 cis region, pre-formatted. All SNPs in LD panel |
| `sjogren_sumstats` | 341 | Raw outcome (case-control) | Sjögren's Disease GWAS, all chromosomes |
| `sjogren_outcome` | 50 | Standardised outcome | SjD chr20 CD40 region (1 real + 49 synthetic). All SNPs in LD panel |
| `cd40_decode_gwas` | 10 | deCODE GWAS | Dummy deCODE format data, input for `format_pqtl_decode()` |
| `cd40_decode_variants` | 10 | deCODE variants | Allele frequencies for `cd40_decode_gwas` |

### LD reference panel

`inst/extdata/ld_ref.{bed,bim,fam}` — PLINK binary fileset of **real 1000
Genomes phase 3 genotypes** (release 20130502, GRCh37) for the 503 EUR
individuals at the 50 `cd40_exposure` SNPs (chr20:44666595–44782428). Total
size ~6 KB. It is generated by `data-raw/ld_ref.R`, which fetches only the CD40
slice of the public chr20 VCF by remote tabix, checks every SNP's position and
alleles against `cd40_exposure`, and writes the fileset with PLINK. Its
Bioconductor dependencies are not in `DESCRIPTION` (`data-raw/` is
build-ignored), and the fetched genotypes are cached under
`tools::R_user_dir("mrpipeline", "cache")`, so re-running needs no network.

Things to know when writing tests against it:

- **The LD is real and strong.** 147 SNP pairs have |r| > 0.8, and
  rs1883832 / rs4810485 are in perfect LD (r = 1), so any instrument set
  containing both makes the GLS weight matrix singular.
- **Clump leniently for cis-region LD tests.** At the default
  `rsq_thresh = 0.001` the CD40 region clumps to a single instrument; the
  `run_mr()` tests use `rsq_thresh = 0.3` (seven instruments, |r| up to 0.53).
- **rs13045469 is tri-allelic in 1000 Genomes** (G/A/C; the exposure has C/G).
  The panel codes it as C vs not-C (A written as G), which is the exact LD of
  the C-allele dosage.
- **A1 is PLINK's default (the minor allele).** The single-SNP test relies on
  rs1883832's A1 being `"T"`; the generator asserts it.

For edge cases a real panel cannot pin down -- a chosen r between a chosen
pair, or an exactly singular pair -- use `make_ld_panel()` in
`tests/testthat/helper-ld-panel.R`, which simulates a small panel on the fly.

Access the bfile prefix in tests:

```r
bfile <- sub("\\.bed$", "", system.file("extdata", "ld_ref.bed", package = "mrpipeline"))
```

Guard integration tests that need plink:

```r
skip_if_not(file.exists(paste0(bfile, ".bed")), "LD reference panel not available")
```

## PR Checklist

Before opening any pull request:

1. `air format .` — auto-format R files (CLI tool)
2. `lintr::lint_package()` — fix any lint warnings
3. `pkgdown::build_site()` — confirm site builds without errors
4. `devtools::check()` — must produce 0 errors, 0 warnings
5. `devtools::test()` — all tests must pass

## Contributing

<!-- TODO: branch naming convention (phase-N/short-description) -->
<!-- TODO: GitHub issue → feature branch → PR to dev workflow -->
